Winding-Up of Fibrin Fibers as a Novel Mechanism of Platelet-Mediated Fiber Compaction

Version 1

Abstract

This study reveals a previously unrecognized mechanism by which platelets retract and compact fibrin fibers. Using a newly developed 2D fiber-retraction assay, we observed an initial “gearwheel” pattern of actin–myosin organization in spread platelets with associated, extracellular fibrin patches that appear to form an initiation complex for fibrin fiber attachment and rearrangement. The final outcome of this process results in platelets surrounded by tightly packed fibrin fibers, mirroring the architecture of platelets and adjacent fibers within a retracted clot. Thus, the observed compaction process might also take place during clot retraction in order to reduce clot volume, stiffen the clot and enhance wound repair. Apart from pulling on fibers like on a rope, platelets actively wind-up fibrin fibers into compact structures, similar to balls of wool. Besides DNA packaging, this represents a new example of a natural fiber compaction mechanism. Using a combination of 3D clot-retraction and 2D fiber-retraction assays, expansion and electron microscopy, live imaging and mathematical modeling, we show that platelets use an actomyosin-driven swirling motion to gather and loop fibrin fibers around the base of bulbous protrusions (“bulbs”). These bulbs form when a platelet becomes trapped between fibrin fibers during clot retraction or 2D fiber-retraction assays. These findings complement and extend earlier models of platelet-mediated fibrin fiber retractions, offering new insight into how platelets mechanically organize fibrin fibers.

Introduction

Platelets are absolutely essential in case of vessel injury. They are not only necessary to initiate hemostasis but also to retract the forming clot thereby pulling together the edges of the ruptured vessel and flattening the clot for improved blood flow. Different approaches to visualize fibrin fibers in whole blood or plasma clots have shown that fibrin fibers represent the main mechanical and structural framework of clots,1,2 occupying a substantial portion of the clot volume. Previous studies3–6 have shown that platelets are able to retract an unconstrained plasma clot to a tiny volume. The extent and rate of clot retraction are influenced by several factors, including platelet and red blood cell counts, fibrin and thrombin concentrations, and the geometric context of clot formation.7 However, platelets are the primary cellular agents driving such extensive clot compaction, although erythrocytes also contribute.8 How do platelets accomplish this tremendous task? Two key studies offer insight: While Fletcher’s lab demonstrated that platelets generate contractile forces comparable to muscle cells,9 Weisel’s group provided mechanistic insight, showing that platelets extend filopodia to bind and pull on fibrin fibers, much like tugging on ropes10 causing localized fiber accumulation.

However, pulling alone may not fully account for the dramatic reduction in clot volume. We hypothesized that efficient clot retraction requires substantial densification of the fibrin fiber network. Using expansion microscopy, we observed that platelets wind-up fibrin fibers similar to balls of wool around individual bulbs. Using a “2D fiber-retraction assay” we found that fibrin patches are initially arranged in a ring-like pattern in close proximity to the circular organization of the actin cytoskeleton in spread platelets. Over time, wound-up fibrin fibers accumulate above the plasma membrane of platelets in a myosin dependent way.

To explore the mechanistical aspects of the platelet mediated fiber compaction we used computational modeling and live-cell video microscopy. Our results suggest that actomyosin-driven swirling within spread platelets on a 2D surface or within individual bulbs of platelets in the 3D clot space may underlie the observed fibrin coiling and compaction.

Results

Platelets become labeled by fluorescent fibrin fibers during clot retraction

We first estimated the capacity of platelets to retract diluted plasma clots (12% plasma) under unconstrained conditions (Fig. 1A). Under these conditions, 4×107 platelets are able to retract the clot volume to approximately 1.5% of its initial volume (400 μl) within 60 min. Even platelets kept overnight at room temperature before the retraction assay were still retracting the clot to the same extent, however after a longer lag phase. We then used live imaging to follow the reorganization of fibrin fibers during unconstrained clot retraction. Platelet rich plasma (PRP) was spiked with fibrinogen-Alexa 488 before thrombin induced clot formation and image acquisition. Very rapidly one can observe the formation of fluorescent fibrin nodes during the retraction process. Since under the assay conditions the clot is composed only of platelets and fibrin fibers, we hypothesized that the nodes may correspond to platelets surrounded by fibrin fibers (Fig. 1B and video 1). To test this hypothesis, we performed a time course analysis of clot retraction in presence of fluorescent fibrinogen under constrained conditions preventing complete retraction. This approach allowed us to stain and visualize individual platelets in a clot even after prolonged retraction times. Hemostatic clots that form in vivo following vessel injury are also likely to experience mechanical constraints, as they remain anchored to the damaged vessel walls while spanning the breach. Retracted clots were fixed and stained for the αIIb-integrin subunit to detect the localization of platelets within the clot. At the position of each platelet, we consistently observe colocalization of a fibrin node at all time points examined (Fig. 1C), confirming our hypothesis that the fibrin nodes observed in figure 1B form at platelet sites. Furthermore, the results show that the build-up of fibrin fibers around individual platelets occurs very rapidly, within 10 min. The fibrin fibers forming the nodes did not decorate the entire platelet surface but surrounded the center of platelets embedded in the clot. Please note, that platelets occupy a larger space after longer retraction times, but the fiber nodes persist even after 80 min of retraction.

Platelets become labelled by fluorescent fibrin fibers during clot retraction.

A) Time course of unconstrained clot retraction by platelets (4×107 in 400 μl of 12% plasma in PBS) for a total of 60 minutes at room temperature. Experiments have been performed 6x using blood from different donors and typical examples of blood from two different donors are shown (experiments III and IV are repetitions of I and II the day after; although a longer lag-phase is observed, the final retraction volume is similar for the four experiments). B) Unconstrained clot retraction (1×108 platelets per ml in 50% plasma/50% PBS) in presence of fibrinogen-Alexa 488. Image acquisition was started immediately after thrombin addition at a focal plane 100 μm above the bottom of the well and image stacks were collected (61 focal planes, step size 0.5 μm) for a time period of 20:27 min (100 frames with a time interval of 12.5 sec). See also associated Video 1. The experiment has been performed twice using platelets of the same donor. Shown are the first, intermediate and last time points of a depth color-coded time-lapse video (left panel; scale bar 10 μm) and the maximal intensity projections (MIPs) of the same time points (right panel; scale bar 10 μm). C) Constrained clot retraction (1×108 platelets per ml in 50% plasma/50% PBS, fibrinogen-Alexa 488 final concentration 12.5 μg/ml) between two holders. Clots were induced by addition of thrombin (2.5U/ml final), fixed at the indicated retraction times, embedded in gelatin, flash frozen and cryosections (14 μm) were stained for the integrin subunit αIIb (magenta; scale bar 5 μm). The time course was performed twice using PRP from two different donors (retraction assays were repeated, although not for all time points, more than eight times using blood from different donors, with consistent results). Image acquisition was performed using a wide-field epi fluorescence microscope (BX41; Olympus) equipped with a Plan 100x/1.25 NA oil objective, a camera (DP70; Olympus), and the acquisition software analySIS (Olympus).

To gain a better insight about the organization of fibrin fibers around platelets within a clot, we used 4x isotropic expansion microscopy. We found that fibrin fibers form a cage-like structure around the platelet center, clearly visible when scrolling through the image stacks (Fig. 2 A-D and animation, Video 2). This pattern was consistently observed across all platelets in the clot and after different clot retraction times (not shown). Most platelets had formed large bulbs, thin filopodia, or both with fibrin fibers typically localized around the base of the bulbs or along filopodia. It remains unclear whether all fibrin segments reside on the external platelet membrane or whether some of them extend into channels of the open canalicular system (OCS). Some of these channels persist in activated platelets.11 Please note that straight fibers radiate from the platelet, indicating a high tension in the platelet clot environment. We want to emphasize that changes in temperature, or plasma/platelet concentrations change the kinetics of retraction but we always observe the cage-like fibrin fiber organization around individual platelets (not shown).

Fibrin fibers are organized in a “cage-like” fashion around platelets within a constrained clot.

A-D: Two typical examples (out of 39 acquisitions of four experiments using blood from different donors) of platelets in a constrained clot with attached fibrin fibers are shown (1×107 platelets per ml PBS/50% plasma and fibrinogen-Alexa 488). Clot retraction was allowed to take place in an inoculation loop (see methods) for 15 min before fixation and immunofluorescence staining of the integrin subunit αIIb (magenta). Samples were then processed for expansion (scale bars 10 μm = 2.5 μm after correction for expansion; indicated z-levels are not corrected for expansion; see also associated animation, Video 2). A) Four focal planes from a stack of images (z-level as indicated) showing a platelet in a clot (fibrin fibers in green, plasma membrane staining using an antibody against the integrin subunit αIIb in magenta and the merge). B) 3D reconstruction of the image stack used in A (46 planes, step size 0.5 μm), a bulb is indicated by an arrow. C) Four focal planes show another typical platelet in a clot (z-level as indicated). D) 3D image reconstruction of the image stack used in C (28 planes, step size 1 μm), a bulb is indicated by an arrow.

Platelets, spread on a 2D surface, organize fibrin fibers above them

So far, our results demonstrate that platelets possess a remarkable capacity to precisely organize fibrin fibers around themselves when embedded in a clot. To investigate the mechanism by which platelets arrange the fibers around their center we sought to use a model system with reduced complexity. We, therefore, developed a “2D fiber-retraction assay”, which is a reductionist approach that does not replicate full clot retraction, but can provide insight into how platelets interact with and organize fibrin fibers. To this end, a diluted suspension of PRP (2.5×106 platelets/ml, 4μl plasma/ml PBS) was incubated with fluorescent fibrinogen and thrombin for 15 min to induce platelet activation and fibrin polymerization. Platelets were then allowed to spread on a 2D surface for 30 min before fixation. Under these conditions, we observe four categories of spread platelets and associated fibrin fiber organizations. Platelets without fibrin fiber contact (28.3+2%) and platelets with a small fibrin dot at their center, which could be a fibrin initiation complex (40+3%). Another category of platelets had an accumulation of coiled fibers above them (fiber winding 18.3+0.5%) and a fourth category showed strong fiber densification around them (fiber compaction 13.6+0.9%) (Fig. 3, left image of A, quantification B). We then investigated, whether these fiber accumulations depend on myosin actions as it is the case for clot retraction. Myosin inhibition using blebbistatin strongly diminished the build-up of fibers around the spread platelets (fiber-winding 6.6+0.9% and fiber compaction 0.2+0.2%), while a higher percentage of platelets with fiber initiations or fiber contacts was observed (63.4+2%) (Fig. 3, right image of A and quantification B).

Spread platelets accumulate fibrin fibers above them in a myosin dependent way.

A) Shown is a 2D fiber-retraction assay in the presence of DMSO (control) or 50 μM blebbistatin for myosin inhibition. After a 30 min preincubation of platelets (4 μl plasma per ml PBS, see methods section) with or without blebbistatin, thrombin is added and the assay is continued as described in the methods section. Scale bar 10 μm, arrowheads indicate platelets without fiber contacts, thin arrows with fiber initiation or contact, larger arrows indicate platelets starting to wind-up fibers and even larger arrows point to platelets with compacted fibers around them (see legend below bar graph in B). B) Quantification of the percentage of platelets in the different categories using image acquisitions of four experiments using blood from different donors (for control conditions a total of 636 platelets were counted and for blebbistatin conditions a total of 524). Shown are means + sem, evaluation of significance using the Student’s t-test.

Then, we used the 2D fiber-retraction assay followed by expansion microscopy to investigate in more detail how the fibers above spread platelets are organized. Platelets were stained either for the αIIb integrin subunit or for myosin. We observed that fibrin fibers are wound-up above platelets (category “fiber winding” in fig. 3), frequently forming ball-like structures wrapped around a bulbous protrusion in the middle of the spread platelet, often referred to as pseudo-nucleus in the literature (Fig. 4A-E and animation, Video 3).

Coiled-up fibrin fibers above spread platelets.

Two representative examples (out of 18 acquisitions from three independent experiments using blood from different donors) of fibrin fiber accumulations above spread platelets in the 2D fiber-retraction assay (4 μl plasma per ml PBS, see methods section) are shown (see also animation, Video 3). Samples were stained for the αIIb integrin subunit and expanded (scale bars 10 μm = 2.5 μm after correction for expansion, indicated z-levels are not corrected for expansion). A) Three different focal planes (z-levels as indicated) from an image stack showing two spread platelets and attached fibrin fibers. B) 3D reconstruction of the image stack used in A (51 planes, step size 0.25 μm), shown are three different view angles. C) Three different focal planes from an image stack illustrating a spread platelet and attached, coiled-up fibrin fibers. D) 3D reconstruction of the image stack used in C (50 planes, step size 0.25 μm), shown are three different view angles. The fibrin fibers are additionally displayed as depth-color coded (Fire LUT) reconstructions to highlight the straight fiber parts along the z-axis. E) Same fiber image stacks as in C and D. Left panels are MIPs of the green fibrin fibers along the z, x and y axes. Right panels are the same projections color coded in the HSB space (Hue, Saturation, Brightness) according to the orientation of individual fiber segments. The fibers are well aligned along the z-axis, while they crisscross horizontally.

In addition to these spread platelets with coiled fibrin fibers above them, we also observed platelets entangled by many fibers and therefore unable to spread completely. These platelets had formed filopodia and bulbs similar to platelets within a clot and a dense fiber accumulation is observed around the base of their bulbs (category “fiber compaction” in fig. 3A). These fibers appeared to get coiled into multiple loops, tightly compacted around the base of platelet bulbs similar to individual balls of wool (six small “balls” can be distinguished in Fig. 5A, focal planes at 4 and 5.25 μm and animation, Video 4, see also Fig. 6 and animation, Video 5). Winding-up of fibers by platelets implies that fibers around individual platelets may get twisted and thereby bundled. Local twisting of fibers near platelets is indeed observed in most of the presented images, although the resolution of optical microscopy is often insufficient to resolve regions of densely packed fibers. Particularly clear examples are shown in Figure 6. It should be noted that fiber bundles between two platelets can also be twisted (Fig. 6C-E). The enhanced fibrin labeling observed around these platelets, compared to those in a typical constrained clot (Fig. 2A-D), may be due to multiple winding-up of free, unconstrained fibrin fibers.

Strongly compacted fibrin fibers around the center of platelets in the 2D fiber-retraction assay.

Two representative examples of platelets are shown with bulbs encircled by fibrin fibers in the 2D fiber-retraction assay (4 μl plasma/ml, see methods section) behaving similar to platelets in a clot (see also animation, Video 4). Samples were stained for the αIIb integrin subunit (A, B) or the myosin light chain (MLC; C,D) and processed for expansion (scale bars 10 μm = 2.5 μm after correction for expansion, indicated z-levels are not corrected for expansion). Experiment was repeated 4x with blood from different donors with a total of 18 image stacks acquired for fibrin/integrin and 62 for fibrin/myosin staining. A) Three different focal planes from an image stack showing two platelets, one spread and one with bulbs and attached, compacted fibrin fibers. B) 3D image reconstruction of the image stack used in A (34 planes, step size 0.25 μm), shown are three different view angles. C) Three different focal planes from an image stack illustrating another platelet with bulbs and attached, rolled-up, compacted fibrin fibers. D) 3D image reconstruction of the image stack used in C (55 planes, step size 0.25 μm), shown are three different view angles.

Fibrin fibers are twisted in the proximity of platelets.

Two representative examples of platelets with compacted fibrin fibers are shown in the 2D fiber-retraction assay (4 μl plasma/ml, see also animation, Video 5). Samples were stained for myosin II (A, B) or phosphorylated myosin light chain (p-MLC; C-E) and processed for expansion (scale bars 10 μm = 2.5 μm after correction for expansion, indicated z-levels are not corrected for expansion). Please note also the strong fiber accumulation around the platelets similar to platelets shown in figure 5. A) 3D image reconstruction of an image stack (72 planes, step size 0.25 μm) illustrating a platelet with bulbs and wound-up compacted fibrin fibers. B) Three different focal planes (z-level as indicated) from the image stack used in A, depth color-coded. Rectangles indicate twisted fibrin fibers and a corresponding zoom is shown below. C) Three different focal planes (z-level as indicated) from an image stack illustrating two platelets with bulbs and attached, wound-up, compacted fibrin fibers. Twisted thick fiber bundles are observed between the two platelets. D) 3D image reconstruction of the image stack (91 planes, step size 0.25 μm) used in C, shown are three different view angles. E) The same focal planes as in C and the top view of the 3D image reconstruction shown in D of the fibrin fibers, depth color-coded to better distinguish the twist of the thick fiber bundles between the two platelets.

A ring-like organization of fibrin fiber patches can be observed at the center of spread platelets

We next increased the plasma volume (7μl/ml instead of 4μl/ml) in order to enhance the amount of formed fibrin fibers. This modification led to the formation of a fragile fibrin gel, which was partially lost during the fixation or staining steps. As a result, only very few fibers remained attached to the platelets or to the coverslip.

The residual fibrin fluorescence revealed a ring-like arrangement of fibrin patches at the platelet center. This fibrin “rosette” co-localized with a circular actin organization, forming a gearwheel-like pattern with a compacted fibrin mass at the inner edge of the circle. The actin cytoskeleton was either composed of radial and transverse fibers (Fig. 7 A,B and animation, Video 6) or radial fibers and actin nodules (Fig. 7C and animation, Video 6) depending on the degree of platelet spreading.12,13 A faint rosette with a compact fibrin mass and an attached fibrin fiber is shown in figure 7D and animation, Video 6).

A fibrin rosette is located close to a ring-like actin organization in spread platelets.

Four representative examples (out of 20 acquisitions, experiment repeated 4x using blood from different donors) of a fibrin rosette associated with spread platelets in the 2D fiber-retraction assay (7 μl plasma per ml PBS, see methods section) are shown (see also animation, Video 6). Samples were stained for the αIIb integrin subunit as well as for actin and processed for expansion (scale bars 10 μm = 2.5 μm after correction for expansion). A) 3D image reconstruction of an image stack (30 planes, step size 0.33 μm) of fibrin fibers (green) and integrin or actin staining (upper and lower panels respectively, magenta). B) Similar example as in A (17 planes, step size 0.33 μm). C) Another example of a fibrin rosette with intercalated actin nodules (34 planes, step size 0.33 μm). D) An example showing a long fibrin fiber associated with the fibrin rosette (35 planes, step size 0.38 μm).

We then investigated the localization of myosin II under these conditions and found that it is distributed throughout the platelet. However, a denser ring-like organization of myosin is present in the center of the spread platelet, closely associated with the fibrin rosette (Fig. 8 and animation, Video 7). Some myosin spots are intercalated with compacted fibrin (Fig.8 A, focal plane at z=7.8 µm), while wound-up fibrin fibers can be observed at the periphery and above the myosin ring structure (Fig. 8A-F). The intensity of fibrin rosettes is highly variable between platelets of different donors. Intense rosettes as shown in figure 7 were observed for platelets from two donors (approximately 3-5% of spread platelets with fiber initiation complexes), while faint rosettes (as in Fig. 8AB) were observed for 50% of spread platelets with initiation complexes using blood from three other donors.

The fibrin rosette and attached fibers are in close proximity to myosin II in spread platelets

Three representative examples (out of 62 acquisitions, experiment repeated 4x using blood from different donors) of fibrin and myosin localization in spread platelets (see also animation, Video 7) in the 2D fiber-retraction assay (4 μl plasma/ml for A, B, E, F or 7 μl plasma/ml C, D; see methods section). Samples were stained for myosin (A, B, C, D) or phosphorylated myosin light chain (p-MLC; E, F) and expanded (scale bars 10 μm = 2.5 μm after correction for expansion, indicated z-levels are not corrected for expansion). A) Three different focal planes (z-levels as indicated) from an image stack showing a spread platelet and an attached fibrin fiber in the center of a faint fibrin rosette. B) 3D image reconstruction of the image stack used in A (36 planes, step size 0.3 μm), shown are three different view angles. C) Three different focal planes from an image stack illustrating a spread platelet with a fibrin rosette and an attached, rolled-up fibrin fiber. D) 3D image reconstruction of the image stack used in C (31 planes, step size 0.33 μm), shown are three different view angles. E) Three different focal planes from an image stack illustrating a spread platelet with attached, rolled-up fibrin fibers. F) 3D image reconstruction of the image stack used in E (54 planes, step size 0.25 μm), shown are three different view angles.

Modeling platelet-mediated winding-up of fibrin fibers

To address the mechanical aspects of the observed platelet-mediated fibrin fiber organizations, we developed a model based on our experimental observations. We first quantified fiber curvature and the number of platelet extensions. Each platelet (in clots or on a glass surface surrounded by fibers and therefore unable to spread) showed 7-10 bulbs/filopodia (Fig. 9A-C). Fiber curvature was determined using transmission electron microscopy (TEM) and fluorescent (FLUO) images. Curvatures were similar for both image types (mean values for κTEM=10.09 ± 6.6 μm-1 and κFLUO=6.77 ± 4.7μm1) with differences likely due to the higher resolution of TEM (Fig. 9D-F).

Determination of platelet protrusions and fibrin fiber curvatures in proximity of platelets

A) Three examples of platelets within a clot stained for the aIIb integrin subunit (including platelets shown in figure 2). MIPs are shown, the number of platelet extensions is indicated and has been determined manually by scrolling through the image z-stack. Scale bars 10 μm = 2.5 μm after correction for expansion. B) Three examples of platelets attached to a glass surface (including platelets shown in figure 5) and stained for the MLC (upper panel) or the aIIb integrin subunit (middle and lower panel). MIPs are shown, the number of platelet extensions is indicated and has been determined manually by scrolling through the image z-stack. Scale bars 10 μm = 2.5 μm after correction for expansion. C) The comparison of all data points between platelets in clots or attached to a glass surface does not show a statistically significant difference between protrusions of platelets in a clot or when spread on a glass surface (Mann-Whitney U test). D-F: The degree of fiber curvature in examples shown in D and E is determined as κ=1/r using the radius of a circle fitting the region of strongest curvature of the fibers. D) Three examples of TEM images used to determine the curvature of fibrin fibers close to platelets within a clot (measured examples are indicated by arrows). E) Three examples of fluorescent images of fibers wound around platelets attached to a glass surface (including platelets shown in figure 5). A single focal plane of each image stack is shown (scale bars 10 μm = 2.5 μm after correction for expansion). F) Comparison of all data points obtained using TEM images and fluorescent images (curvature in κ μm-1). For quantification, 31 EM acquisitions of three clots from different donors were used as well as 12 Fluo acquisitions of platelets from 4 different donors.

Our model incorporates these observations, published parameters (Table 1, methods section) and builds on the framework established by Bershadsky and colleagues.14 They demonstrated that in fibroblasts, cultured on circular substrates, the actin cytoskeleton is composed of radial actin fibers extending towards the cell center and circular transverse fibers enriched in myosin-IIA. Live-cell imaging in their study revealed that the transverse fibers slide centripetally along the radial fibers, generating a chiral swirling motion directed towards the cell center. Fibronectin-coated beads bound to integrins on the cell surface are dragged along with the intracellular swirling of the cytoskeleton suggesting that extracellularly bound fibrin fibers may experience a similar rotational motion. Chiral rotations of the cytoskeleton are not only observed at the level of entire cells, but also in subcellular regions, such as individual filopodia of the neuronal growth cone.15,16

Along this line, our simulations suggest that fibrin fibers are anchored to integrins in the plasma membrane of bulbs and filopodia, are pulled towards the platelet center by retraction of filopodia, as demonstrated by Kim et al.10 Spiral cytoskeletal motions within each bulb drive integrins, and thus the attached fibrin fibers, towards the platelet center, wrapping the fibers around the base of the bulbs (Fig. 10A and animation, Video 8). The efficiency of loop formation depends on the initial fiber position (Fig. 10B-D). The process effectively compacts long fibers into loops with a radius of ∼420-nm near the base of bulbs (Fig. 10D).

Simulation of platelet mediated winding-up of fibrin fibers.

A) Time evolution of the fibrin fiber configuration around the platelet bulb. Two different angles between the fiber and the x-axis were used ([1]: α = π/4, lime color; [2]: α = π/2.5, dark green color). Left: initial positions, top view and side view. Center: intermediate position, top view and side view. The fibrin fiber began to form a loop around the bulb. Right: final position, top view and side view. A compact fibrin loop was formed around the bottom of the bulb (see also animation, Video 8). B) Time when the first node bound to the bulb and thus the formation of the fibrin loop started. The fibrin winding-up started earlier for a larger angle between fiber and x-axis (and consequently smaller angle between fiber and tangent line at the point of the node 0 binding). The number of nodes that were bound to the bulb at the end of the simulation was slightly higher for larger α. Thus, larger α led to a more compact fibrin fiber organization. C) The distance between nodes and the bottom of bulb along z-axis. The distance was smaller for larger α and the fibrin organization was slightly more compact. D) Comparison of final fiber configurations around the bulb for two different α.

Transmission electron microscopy (TEM) provides evidence of curved and cross-sectioned fibrin fibers at the base of platelet bulbs

To confront the results obtained by expansion microscopy and model simulations with those using TEM, we analyzed TEM images of platelets and fibrin fibers within a clot. We observed numerous instances of cross-sectioned fibrin fibers at the base of bulbs as illustrated in Figure 11. This figure also shows two examples suggesting fibers that curve along bulbs from top to bottom and fibers attached to a filopodium.

Features predicted by the model/simulation are observed on TEM images.

Interpreted, potential fiber organisations are highlighted in green on an identical image on the right. A) Fibrin fibers appear to encircle a bulb. B) Cross sectioned fibrin fibers at the base of a bulb, which potentially curve around the bulb. C) Several cross sectioned fibers at the base of a bulb, which may surround the bulb. D) Several cross sectioned fibers along a platelet extension potentially forming a spiral. E) Cross sectioned fibrin fibers around a forming bulb. F) Cross sectioned fibers at the base of a bulb and two fibers along a filopodial extension.

Live imaging of platelet-mediated reorganization and compaction of fibrin fibers

We performed live video microscopy using platelets from three different donors to observe how platelets wind-up fibrin fibers (Fig. 12 and Video 9). In many cases, fibrin fibers were already accumulated above platelets at the start of imaging, with minimal further fiber movements or compactions; likely due to fibers binding to each other, to the dish or to the platelet surface (Fig. 12A). However, dynamic fiber movements, both clockwise and counterclockwise, were observed above two platelets present within the rectangle indicated in figure 12A (see zoom images lower panel). An arrow in Fig. 12A marks the formation of a kink in a fibrin fiber. For the second experiment (blood from a different donor), we reduced the incubation temperature to 21°C in order to slow down the platelet and fiber movements. In the rectangle indicated in figure 12B, a fibrin fiber gets progressively curved and eventually ruptures (at time point 00:48). The resulting free end of the fiber then undergoes a clockwise rotation (see zoom images lower panel). Another platelet producing a fiber densification is marked by an arrow in figure 12B. The video shows a counterclockwise rotation of the central part of the platelet during the fiber densification process. A third experiment (platelets from another donor), again performed at room temperature, shows the rotational (counterclockwise) movement of the pseudo-nucleus and alongside the winding-up of fibers around this bulbous protrusion. This apparently generates strong tension that pulls the pseudo-nucleus towards the edge of the spread platelet, where the movement ceases (Fig. 12C, see zoom images lower panel). These fibrin fiber movements occurred without visible filopodia-fiber interactions, suggesting that the bound fibers are dragged by intracellular movements of the cytoskeleton. Since still images cannot fully illustrate the fiber movements or the rotation of the platelet center, we invite the reader to consult the accompanying time-lapse videos (Video 9). After the time-lapse acquisitions, samples were fixed and stained for the αIIb-integrin subunit to localize the position of platelets on the dish surface. The final fibrin fiber organization around platelets in these live videos of the 2D fiber-retraction assays closely resembles the arrangement of fibers around platelets in a constrained clot, except that some platelets are completely spread on the surface with no or very few fiber contacts. Platelets surrounded by densely packed fibers had formed bulbs and filopodia, but remained attached to the dish surface (Fig. 12D).

Live imaging of spread platelets winding-up fluorescent fibrin fibers.

A-C: Shown are three experiments using blood from different donors. A) Washed platelets were adjusted to 5×106 platelets per 2 ml PBS and 5 μl plasma was added as well as fibrinogen-Alexa 488. Thrombin was added to induce fibrin fiber formation and platelet activation. After 10 min, the suspension was transferred into a petri dish (WPI Fluoro), centrifuged and installed in the microscope incubator at 37°C to start image acquisition (see also Video 9, upper panel). The first and last time points of the time-lapse video are shown; acquired using the fluorescence and transmission channels. Projections of four focal fluorescent image planes of fibrin fibers (green) and projections of the transmission planes as well as the merged channels are shown. An arrow indicates a forming kink in a fibrin fiber (scale bar 10 μm). Below are shown Individual time points of the time-lapse video zooming on the rectangle indicated in the panels above. This region shows rotational fiber movements. B) Independent experiment under conditions as described in A except that the petri dish was kept at room temperature to slow down the platelet mediated fiber reorganizations (scale bar 10 μm). An arrow shows platelet mediated fiber compaction. Below are shown Individual time points of the time-lapse video zooming on the rectangle indicated in the panels above. This region shows a fibrin fiber getting curved by the platelet and finally ruptures followed by a rotational movement of the ruptured fiber. C) Platelets of a third donor (not washed, i.e., 5 μl PRP with 4×106 platelets per 2 ml PBS) and fibrinogen-Alexa 488 were added to 2 ml of PBS and the experiment was continued as in B (scale bar 10 μm). Below are shown Individual time points of the time-lapse video zooming on the rectangle indicated in the panels above. This region shows fibrin fibers getting coiled around the pseudo-nucleus of the spread platelet. The last zoomed image shows a temporal color-coded projection of the transmission channel to illustrate the rotational movement of the pseudo-nucleus (scale bar 5 μm). D) Samples shown in B and C were fixed 3h after the start of time-lapse acquisitions and stained for the integrin subunit aIIb. Shown are fibrin fibers (green) integrin staining (magenta) and the merge (samples B and C, upper and lower panels, respectively; scale bar 10 μm). E) Preformed fluorescent fibrin fibers were prepared in the absence of platelets. PRP was adjusted to 2.5×106 platelets per ml with PBS and 1800 μl was transferred into a petri dish (WPI Fluoro). The dish was centrifuged to allow spreading of the platelets and the supernatant was replaced by the preformed fibrin fibers. The dish was installed in the microscope incubator to start image acquisition (see also Video 9, last panel). The experiment was performed twice with blood of the same donor. The last fluorescent image of the time-lapse video is shown (left image, scale bar 10 μm). Fifteen of the fibrin accumulations in the middle of the platelets are rotating in a counterclockwise direction (indicated as “1”), four rotate clockwise (“2”) and for 16 no clearly visible turn is observed (“3”). A projection of all fluorescent time points, temporal color-coded using the LUT spectral, is shown. White corresponds to the sum of all the colors at each time point, meaning no movement during the time-lapse video. (right image, scale bar 10 μm).

To minimize pre-imaging fiber buildup, we next added preformed fibrin fibers to pre-spread platelets and started imaging immediately. Under these conditions, surface coating is observed and only a few fibers are present, yet a central fibrin accumulation is consistently observed on each platelet, possibly representing a nucleation site. Among 35 platelets analyzed, 15 displayed a central fibrin mass rotating counterclockwise (viewed from below), 4 rotating clockwise, and 16 showed no consistent directional movement (Fig. 12E). In some instances, the rotating complexes stalled and stopped moving. These observations support the idea that cytoskeletal swirling drives fibrin reorganization above spread platelets.

Discussion

Using a 2D fiber-retraction assay, we identified a previously unrecognized capacity of platelets to compact fibrin fibers into a small volume by winding them into coiled structures during platelet induced fiber retractions. To our knowledge the only example of natural fiber compaction is coiling of DNA molecules.17 In contrast to intracellular, enzyme-mediated DNA compaction, extracellular fibrin fiber compaction requires the cellular activity of platelets. Our investigation began with the observation that platelets become labeled with fluorescent fibrinogen during clot retraction as previously shown by Brzoska et al.18 They demonstrated that fibrin binding to thrombin activated platelets depends on αIIbβ3 integrins and the actin cytoskeleton. While fibrin fiber retraction by platelets in a clot has been attributed to non-muscle myosin II,19,20 the mechanism underlying fiber organization and compaction has remained unclear.

One key mechanism, revealed in previous studies, is the extension of platelet filopodia that bind to fibrin fibers and pull them inwards in a hand-over-hand motion.10 This is clearly the main process by which platelets retract fibrin fibers in a clot. However, this mechanism alone cannot explain the cage-like fiber organization observed around platelets in a constrained clot (Fig. 2). Our findings introduce an additional, distinct mechanism by which platelets can organize fibrin fibers. Platelets wind-up fibrin fibers into dense, ball-like structures, which is an efficient form of compaction (Fig. 3-6). Thus, our model reveals an additional mechanistic aspect of platelet mediated fiber retraction beyond those described by Michael et al.21 and Kliuchnikov et al.,22 so far observed only in the 2D fiber-retraction assay. Nevertheless, the platelet morphology and the organization of fibers around them after the retraction process closely resemble those of platelets in a clot and both, 2D and 3D fiber retractions, depend on myosin actions.

The 2D fiber-retraction assay may recapitulate early post-injury events, such as activation of platelets in suspension triggered by agonists released from neighboring platelets, ultimately leading to the spreading of activated platelets onto the subendothelial matrix to anchor the developing clot at the site of vessel injury. This assay revealed two key observations. First, when many fibers are present, some platelets spread with the typical fried-egg phenotype and coil-up fibers above their pseudo-nucleus (Fig. 3, 4). Others form large bulbs in several directions, similar to platelets in a clot, and compacted fibers are observed around the base of each bulb (Fig. 5, 6). Under these unconstrained 2D conditions, fiber accumulation around the bulbs is much higher than in the case of a constrained 3D clot (Fig. 2), where internal tensions limit fiber winding.

Second, when fewer fibers are retained on the culture surface, several spread platelets display a fibrin rosette around their pseudo-nucleus (Fig. 7,8) that colocalizes with actin, myosin and integrins. A dense fibrin mass is seen at the inner periphery of this rosette, sometimes with an attached fiber. This gearwheel-like architecture may represent an early priming/nucleation step of platelet-induced winding of fibrin fibers. The actin cytoskeleton in platelets exhibits a chiral organization23 which, combined with the action of myosin, could be responsible for the swirling motion observed in spread platelets (Fig.12 E-G and Video 9). Integrin-bound extracellular fibers are pulled along by this vortex inducing their winding-up above the pseudo-nucleus (Fig. 12C). Fibrin rosettes may fade as coiled fibers become more prominent, suggesting a transition from an initiation phase to active coiling and compaction.

Ideas and Speculations

How this 2D “gearwheel” translates to the 3D clot environment remains unclear. One possibility is that the pseudo-nucleus may correspond to a bulb around which fibers are coiled and accumulate. Platelets with bulbs within a clot have been described previously.24–27

We wondered whether platelet bulbs might represent plasma membrane blebs similar to membrane protrusions formed by megakaryocytes during clot retraction.28 However, the consistent presence of 8–9 bulbs per platelet (Fig. 9A-C) suggests that they reflect a “three-dimensional spreading” morphology, reminiscent of the eight corners of a cube, allowing platelets to sense the clot’s 3D environment along the main directions. Unlike blebs, these bulbs often contain organelles such as mitochondria (Fig. 9D, upper image and Fig. 11B) implying that each bulb may have its own energy supply and actomyosin organization. This could enable swirling within each bulb and explain the observed fibrin fiber accumulation at their base. Thus, at the onset of clot retraction, each emerging bulb may act as a discrete functional unit onto which the gearwheel assembles, enabling fibrin fibers to attach and initiating their subsequent winding and compaction (see scheme, Fig. 13). Indeed, blebs of megakaryocytes retracting a clot may behave similarly as platelet bulbs since fibrin fibers can be seen surrounding the base of blebs during megakaryocyte-mediated clot retraction, as can be observed in the study by Kim et al. (see the supplementary movie S3 of this publication).28

Schematic illustration of the hypothesis explaining the potential mechanism by which platelets may wind-up fibrin fibers and compact them.

Represented is a platelet in the 2D fiber-retraction assay (upper part) or in a clot (lower part). At the beginning of platelet fiber interactions, a fibrin rosette is present around the pseudo-nucleus and in each bulb with a dense fiber mass at the internal periphery of the rosette possibly representing an initiation complex. Radial actin fibers extend to the center of the platelet and an actomyosin rotational movement along the radial actin fibers may lead to the winding-up and the downward sliding of the extracellular attached fibers to encircle the pseudo-nucleus or the base of the bulbs at the final state of compaction (3D rendering of the scheme was produced using an open AI).

Our live time-lapse videos of platelet-mediated fiber movements in a 2D environment support the concept of a cytoskeletal vortex that drives the coiling and compaction of extracellular attached fibers. Additionally, the other time-lapse video capturing the rotation of compacted fibrin at the platelet center further suggests the presence of cytoskeletal swirling within spread platelets. This swirling may play a role in scanning the extracellular space for forming fibers that can be anchored to the initiation complex. The process occurs on a notably slow time scale, which could facilitate effective fiber capture by the rotating complex. However, this rate appears insufficient to account for the rapid labeling of platelets during clot formation, suggesting that regulatory mechanisms, possibly involving mechanical feedback, may be at play. Indeed, as mechanosensory cellular particles, platelets adjust their contractile forces and dynamics in response to environmental cues such as fiber tension and clot stiffness.9,10,29,30,22

In conclusion, our study identifies a novel mechanistic aspect of platelet function: the active winding-up and compaction of fibrin fibers. This complements the existing model of platelet-mediated clot retraction10 and provides a new perspective on how platelets may contribute to clot architecture, mechanical stability, and wound healing.

Materials and methods

Reagents and antibodies

Reagents: Neutral formalin (Sigma-Aldrich, Saint-Quentin Fallavier, France), Pluronic F-127 (Sigma-Aldrich, P2443), thrombin (Sigma-Aldrich, T4648), Jasp-K4-HA (a HAK-actin probe, kind gift of Paul Guichard/Virginie Hamel),31 apyrase (Sigma-Aldrich, A6535), heparin (Sigma-Aldrich, H3393), PGI2 (Sigma-Aldrich, P6188)

Antibodies: mouse anti-integrin αIIb (Santa-Cruz, sc-7310), monoclonal rabbit anti-HA (Cell Signaling, C29F4), rabbit anti-non muscle myosin II heavy chain (Covance, PRB-440P), rabbit anti-phospho S19 myosin light chain (Abcam, ab2480), monoclonal rabbit anti-myosin light chain 2 (Abcam, ab92721), Cy3 goat anti-mouse IgG (Jackson, 115-165-166), Cy3 goat anti-rabbit IgG (Jackson, 111-165-003), Alexa 647 goat anti-rabbit IgG (Jackson, 111-605-003)

Preparation of human platelet rich plasma (PRP) and platelet free plasma (PFP)

Whole blood was obtained from the French blood bank (age-range 18-65, mixed genders; pooled blood from multiple donors has not been used). The IRB has authorised our institute’s research projects using human samples under the number DC-2022-4976. PRP was prepared from whole blood drawn into Na-citrate Vacutainers (3 tubes of 2.7 ml per donor). The blood is pooled into a 15 ml falcon tube and centrifuged for 12 min 400g, RT, no brake. The upper phase is the platelet rich plasma (PRP), in which platelets are distributed in a gradient (low concentration on top, high concentration at the bottom). In order to increase the platelet concentration in the final PRP, the upper 3 ml are removed and used to prepare platelet free plasma (PFP, by centrifugation at 12000g, 1 min, RT). The remaining upper phase is collected as the final PRP and the platelet concentration determined.

Platelet washing procedure

PRP was diluted 1/4 with Tyrode/0.35% BSA buffer and platelet activation inhibitors (0.02 U/ml apyrase, 1 μm PGI2, 10 U/ml Heparin, final concentrations) were added. Samples were then centrifuged for 15 min at 900g at RT. The supernatant was aspirated, and platelets were gently resuspended in fresh Tyrode/BSA buffer, 0.02 U/ml apyrase, 1 μm PGI2. The suspension was centrifuged as before and the platelet pellet was resuspended in fresh Tyrode/BSA buffer without inhibitors.

Clot retraction assays

We used three different clot retraction assays illustrated in Supplementary Figure 1.

Unconstrained clot retraction (Fig. 1AB): See paragraph “Live video microscopy” below for method details.

Constrained clot retraction between two holders (Fig. 1C): In order to perform immunofluorescence stainings of individual platelets within a clot, we used a clot retraction assay around two sterile inoculation loops as previously described.32,33 Briefly, PRP was adjusted to 1×108 platelets per ml in 50% plasma/50% PBS (to reduce fibrin fiber density), fibrinogen-Alexa 488 final concentration 12.5 μg/ml and 4 μl/ml erythrocytes for color contrast. 400 μl of this suspension was used for clot induction by addition of thrombin (2.5U/ml final) and immediately transferred to 2 ml Eppendorf tubes coated with 2% Pluronic® and containing the inoculation loops. After different retraction times, clots were fixed for 1 h with isotonic formalin (9 volumes formalin / 1 volume 10xPBS). Clots were then washed with PBS and incubated ON at 4°C in 1 ml PBS/15% sucrose. Clots were further incubated in 1 ml PBS/15% sucrose/7.5 % gelatin at 37°C for 4 h and then inclusion blocks were formed in the same solution on ice. Blocks were then trimmed and snap frozen for 1 min in isopentan at −60°C and stored at −80°C until 14 μm thick transversal sections were cut using a cryostat.

Constrained clot retraction within an inoculation loop: Small clots of 30 μl platelet suspension (50% plasma/50% PBS; 1×107 platelets/ml; fibrinogen-Alexa 488 final concentration 12.5 μg/ml; 4 μl/ml erythrocytes for color contrast) were induced to form within plastic inoculation loops (4 mm internal diameter, attached to a glass coverslip) by addition of thrombin to a final concentration of 2.5 U/ml as described previously.13 Five minutes after clot induction, clots were covered with 500 μl PBS and kept in a cell culture incubator for 10 min. Clots were then fixed with 500 μl isotonic formalin for 15 min, RT, in the dark and then kept in PBS at 4°C until further use. Forming a clot within a small inoculation loop, allowed us to do expansion microscopy without performing cryosections of larger clots, which are difficult to detach from slides during the expansion process.

2D fiber retraction assay

PRP was diluted in PBS (2.5×106 platelets/ml, plasma concentrations were kept at 4 or 7μl/ml depending on the amount of fibrin fibers to be formed; see explanation below) and fibrinogen-Alexa 488 was added to a final concentration of 12.5 μg/ml. To induce fibrin fiber polymerization and platelet activation, thrombin was added (final concentration of 2.5U/ml). The suspension was gently mixed once (to prevent platelet aggregation) and after 15 min of incubation at RT in the dark, 400 μl was transferred into wells of a 24 well plate containing glass coverslips. The plate was centrifuged 3 min, 600 g, RT to allow synchronised contact of all platelets with the glass surface. The plate was then placed in a cell culture incubator for 30 min to allow platelet spreading. Samples were then fixed with isotonic formalin for 15 min at RT and then kept in PBS at 4°C until further use.

To promote fibrin fiber formation, while preventing the development of a clot, the plasma concentration must be kept low (4 μl plasma per ml PBS). This can be achieved by adding a volume of PRP containing 2.5×106 platelets to 1 ml PBS, provided the platelet concentration of PRP is at least 6.25×108/ml (condition used for Fig. 3, 4, 5, 6, 8ABEF). If the platelet concentration in the PRP is lower, platelets must be washed and the appropriate amount of plasma added separately.

At higher plasma concentrations (7 μl plasma per ml PBS), more fibers polymerize leading to the formation of a fragile clot which is aspirated during fixation/washing procedures. Under these conditions very few fibers are left behind. However, a fibrin rosette can be observed in the centre of several spread platelets (condition used for Fig. 7, 8CD).

The 2D fiber-retraction assay followed by immunofluorescence staining and expansion has been repeated 10x with blood from different donors. A total of 18 image stacks stained for fibrin/integrin, 20 stained for fibrin/integrin/actin and 62 stained for fibrin/myosin have been acquired.

Immunofluorescence

Fixed platelets were permeabilized with PBS/0.2% Triton X-100 for 15 min at RT and then incubated with blocking buffer (3% BSA and 10% goat serum in PBS) for 1h at RT. Coverslips were then incubated for 2 h with a primary antibody diluted in blocking buffer, then washed three times and incubated with a secondary antibody diluted in blocking buffer for 2h at RT. After three washing steps, coverslips were either mounted in Mowiol or processed for expansion.

Expansion microscopy

Expansion microscopy was essentially performed according to Gambarotto et al.34 with minor modifications as described previously33. Briefly, fixed, immunostained platelets or clots on coverslips (12 mm diameter) were incubated in PBS/0.7% formaldehyde/1% acrylamide over night at room temperature. The coverslip was then placed upside down onto an ice cold 35 μl drop of gelation solution (PBS/19% Na-acrylate/10% acrylamide/0.1% bisacrylamide) pipetted onto parafilm on ice immediately after addition of TEMED and ammonium persulfate both to a final concentration of 0.5%. After 5 min of incubation on ice, the samples were transferred to 37°C for 1 h. The coverslip plus gel facing up was then incubated in denaturation buffer (5.7% SDS/0.2M NaCI/50mM Tris, pH9) for 15 min at RT with gentle agitation to detach the gel from the coverslip. The gel was then transferred into an Eppendorf tube with fresh denaturation buffer and boiled for 30 min at 95°C. To expand the gel, it was placed into a large volume of H20 and the water was changed twice before incubation ON at RT in fresh H20 leading to 4x isometric expansion.

Image acquisition

For image acquisition of some of the expanded samples (Fig. 2, 5CD, 6, 7, 8), a confocal microscope ConfoBright (Nikon A1R+MP) equipped with a home-made adaptive optics module was used with a 40x/1.15 long distance water immersion objective. The geometric optical aberrations were corrected both in excitation and detection light paths in the open loop mode with a deformable mirror (AlpAO DM97-15) inserted between the confocal scanning head and the microscope inverted stand (Nikon Ti2E). Local aberrations were sensed using the metrics of molecular brightness, derived from fluorescence fluctuations, and were iteratively corrected as individual Zernike modes. This configuration allowed to compensate for the refractive index mismatch, the specimen tilt and the eventual light scattering in depth of expanded samples and to bring back the confocal resolution to its diffraction limit in the vicinity of the imaged platelet.

Image acquisition of other expanded samples (Fig. 4, 5AB) was performed with a confocal microscope Dynascope (LSM 710 ConfoCor 3; Zeiss), 63×/1.4 NA Plan Apochromat objective and the acquisition software Zen 2010. The same microscope and software were used for live imaging (Fig. 12) using the Apochromat 40×/1.2 NA water-immersion objective. Axial Z-stacks (step 0.45 µm) were imaged in a time series using the avalanche photodiode detector in photon counting mode at low 488 nm excitation power (0.1% transmission) to limit photobleaching and phototoxicity, the 633 nm laser was set to 0.6% in order to obtain the transmission light images of the platelets. The pinhole was closed to 0.25 Airy units (14 µm) to increase the spatial resolution. The acquisition of transmitted light was concomitant to the fluorescence.

The FIJI software was used for image processing and analysis.35 No treatment that modifies the raw images was applied. FIJI software was used for the Maximum Intensity Projection (MIP) images, 3D reconstructions of image stacks and for the depth-color coded projection image using the rainbow LUT (Fig. 4D) as well as for the depth-color coded images fire LUT (Fig. 6) and the temporal-color coded image spectral LUT (Fig.12E).

Transmission electron microscopy (TEM)

Clots were formed around two holders as described previously.32 Briefly, PRP was adjusted with PBS to 10 % plasma, 1×108 platelets/ml and 4ml erythrocytes per ml were added for color contrast. Clot formation was induced by addition of thrombin (final concentration of 2.5 U/ml). After 1h of retraction at 37°C, clots were fixed using 1,5 ml of 2,5 % glutaraldehyde for 1h at RT. After fixation, a contrast-enhancing step was performed by incubating the samples in a solution containing 1.5% potassium ferrocyanide and 1% osmium tetroxide in 0.1 M sodium cacodylate buffer. Following this, the samples were embedded in EPON resin according to the protocol described by Eckly et al.36 Thin sections (100 nm) were cut and stained with uranyl acetate and lead citrate and examined using a Jeol 1200 Plus transmission electron microscope operating at 120 kV.

Live video microscopy

- Unconstrained clot retraction (Fig. 1B):

Thrombin (final concentration of 2.5 U/ml) was added to 0.4 ml of a platelet suspension (50% plasma/50% PBS; 1×108 platelets/ml; fibrinogen-Alexa 488 final concentration 12.5 μg/ml; 4 μl/ml erythrocytes for color contrast) to induce clot formation. After quick and gentle mixing, the suspension was transferred immediately into a well (8-well Lab-Tek, coated 2% Pluronic® to prevent clot attachment to well walls). Acquisition was started immediately at 100 μm above the bottom of the well using the confocal microscope ConfoBright (for microscope details see paragraph image acquisition).

- 2D fiber-retraction assay (Fig. 12A-D):

Platelets (washed Fig. 12AB or not Fig. 12C) were adjusted to 2.5×106 platelets/ml PBS and 2.5 μl plasma was added and fibrinogen-Alexa488 (final concentration of 12.5 μg/ml). 15 minutes after thrombin induced fibrin fiber formation and platelet activation, 1800 μl of the suspension was transferred into a petri dish (WPI Fluoro Dish, FD35–100). The dish was centrifuged 3 min, 600 g, RT and installed in the microscope incubator at 37°C (Fig. 12A) or room temperature (Fig. 12BCD).

- 2D preformed fiber-retraction assay (Fig. 12E):

Preformed fluorescent fibrin fibers were prepared in the absence of platelets (2.5 μl plasma/ml PBS; fibrinogen-Alexa 488, final concentration 12.5 μg/ml; thrombin, final concentration 2.5U/ml; incubation 15 min RT in the dark). PRP was adjusted to 2.5×106 platelets/ml with PBS and 1800 μl was transferred into a petri dish (WPI Fluoro Dish, FD35–100). The dish was centrifuged 3 min, 600 g at RT to allow spreading of the platelets. After centrifugation the supernatant was replaced by the preformed fibrin fiber suspension and the dish was installed in the microscope incubator at 37°C.

For all time-lapse videos, image acquisition was started immediately after adjusting the focal position and image parameters (<1 min) using a confocal laser-scanning microscope (ConfoBright or Dynascope, for microscope details see paragraph image acquisition).

The model of fibrin winding-up

The platelet bulb was modelled as a straight cylinder (the radius R = 400 nm, the length = 1500 nm). We assumed that the internal cytoskeleton of the bulb underwent the swirling motion with the angular velocity Rate_swirl as it was shown in14 and simultaneously moved down to the bottom of the bulb with the velocity Rate. Integrins αIIbβ3 on the bulb surface were associated with internal cytoskeletal proteins,37 thus, they underwent the swirling and downstream motion with the same rates.

The fibrin fiber was modelled as it was done previously21,22 with several simplifications. The fiber composed of N equally spaced nodes connected by elastic springs (see Supplementary Fig. 2A). The spring equilibrium length was L and it was equal to the initial distance between nodes. The fiber was located in the XY-plane and the angle α between fiber and x-axis was variable (Table 1). The motion of the node i in solution was described by the second Newton’s law of motion. The system was assumed to be in an overdamped regime (mä = 0): (1) η⋅r˙i=Fi,i−1+Fi,i+1+FiB where ri was the position vector of the node i in three-dimensional coordinates (x,y,z); η was the drag coefficient for the fiber node in the blood serum; Fi,i−1 = Kr · (Δri,i−1 − L) and Fi,i+1 = Kr · (Δri,i+1 − L) were the vector forces which acted on the node i from the node i-1 or i+1 according to the Hook’s law due to the spring elongation or shortening (Δri,i−1 and Δri,i+1 was the distance between nodes i and i-1 or the nodes i and i+1; Kr was the fiber spring constant); FiB was the Brownian force which acted on the node i due to thermal fluctuations. To solve the equations for the nodes in solution, the system of equations was discretized and the Heun’s method was used. According to this method, the evolution of coordinates in time for the differential equation (2) x˙=f(x,t) was calculated using a two-step algorithm: (3) x~n+1=xn+Δt⋅f(xn,tn) (4) xn+1=xn+Δt2⋅{f(xn,tn)+f(x~n+1,tn+1)} where Δt was the time step used in calculation and discretization scheme; tn+1 = tn + Δt; xn was coordinate at the tn moment of time.

Thus, the evolution of the node i coordinates in time was calculated according to a two-step process (in a vector form): (5) r~in+1=rin+Δtη⋅(Fi,i−1(rin,ri−1n)+Fi,i+1(rin,ri+1n)+Fin,B) (6) rin+1=rin+Δt2η⋅(Fi,i−1(rin,ri−1n)+Fi,i+1(rin,ri+1n)+Fin,B+Fi,i−1(r~in+1,r~i−1n+1)+Fi,i+1(r~in+1,r~i+1n+1)+F~in+1,B)

The Brownian force in three dimensions was calculated as described previously.21 Briefly, the mean square displacement of a particle over a single time step was ⟨x2⟩=6kbTηdt, and the iscretized Brownian force acting on the node i was Fin,B=εi⋅6kbTηdt, where εi was sampled from the standard normal distribution.

At the initial time point, the node 0 was bound to the integrin receptor on the bulb surface (top view, Supplementary Fig. 2B). The distance of the binding point from the bulb top was D. The height of the binding point from the bulb surface was equal to the integrin αIIbβ3 molecule length (H = 20 nm) in its elongated conformation.38 The equations for the node 0 swirling and downstream motion were:

When the node i entered the light-magenta zone (Supplementary Fig. 2B), and its distance from the bulb surface became smaller than H, it bound the integrin molecule and its equations of motion became similar to the node 0 equations (7) – (10).

The integrin αIIbβ3 surface density on the bulb surface was relatively high (the distance between the neighbouring molecules was approximately 15 nm). Given such a high density, we assumed that when the node became close enough to the bulb, it immediately encountered the integrin molecule and bound it. Thus, when the distance between the node i and the bulb surface became smaller than H, node i bound to the integrin molecule and the node’s equations of motion became similar to the node 0 equations (7) – (10). The binding was assumed to be irreversible.

The model parameters were listed in the Table 1 below. The model equations were solved using Python 3.11.

eLife Assessment

This study presents a valuable perspective on platelet-mediated fibrin compaction, proposing that fibrin fibers undergo "winding" or coiling, an intriguing framework with potential implications for thrombosis and clot mechanics. However, the evidence supporting an active platelet-driven winding mechanism remains incomplete, relying largely on correlative observations without direct or quantitative validation of the proposed dynamics. Overall, the work is thought-provoking and of clear interest to the field, but stronger mechanistic evidence will be required to substantiate the central claims.

Reviewer #1 (Public review):

This paper reports a previously unrecognized mechanism by which platelets compact fibrin fibers during clot retraction. Rather than simply pulling on fibers, the authors propose that platelets generate swirling motions that wind and loop fibrin into dense structures.

While the results are intriguing, the underlying physical mechanism remains unexplained. In particular, it is unclear how platelets generate swirling motion capable of inducing fibrin coiling, especially when suspended in 3d fibrin mesh. This raises concerns about the conclusions. Also, does fibrin have inherent chirality or structural asymmetry that could promote coiling independently of platelet activity? Furthermore, platelet retraction typically involves platelet aggregation rather than isolated cells, and it is unclear how fibrin coiling would proceed in clustered platelets.

Reviewer #2 (Public review):

Summary:

Grichine et al. investigate platelet-mediated fibrin compaction using human donor platelets and propose a novel mechanistic model in which platelets generate contractile forces and wind fibrin fibers into compact coiled structures. Using a combination of 2D spread assays, 3D clot imaging via expansion microscopy, live-cell imaging, and computational modelling, the authors present evidence of cage-like fibrin architectures, coiled-fibre morphologies, and platelet-centred "rosette" structures present during fibre compaction. They further suggest that actomyosin-driven cytoskeletal dynamics, potentially involving rotational or swirling motion, underlie this proposed winding mechanism, analogous to DNA looping and compaction. The study addresses an important and longstanding question in thrombosis and hemostasis and offers a conceptually novel perspective on clot compaction.

Strengths:

The integration of multiple imaging modalities is a notable strength of this paper. In particular, the 2D fiber-retraction assay provides a useful model for understanding the spatio-temporal dynamics of platelet-mediated fibrin compaction, which can be applied to other systems and may yield detailed mechanistic insights into biological processes. The live-imaging approaches are particularly well executed and offer valuable dynamic insight.

Weaknesses:

The primary weakness of this paper lies in its descriptive nature and its reliance on correlative rather than causal evidence. Several interpretations are not uniquely supported by the data presented. For example, the categorisation of fibrin accumulation in 2D assays as "fiber winding" and "fibre compaction" remains descriptive without establishing winding as a mechanism. Alternative mechanisms, such as circular bundling, stacked fibers under tension, or fibrin crosslinking-induced aggregation, are neither excluded nor investigated. Although the authors present compelling live imaging, establishing winding as a dynamic phenotype would require quantitative analyses, such as measuring angular velocities and coiling rates. The use of a second fluorophore-labelled fibrin population could further strengthen evidence for rotational dynamics. Similarly, the inference of rotational contractility or actomyosin "swirling", based on chiral actin organisation and blebbistatin treatment, is not sufficiently supported to conclude that platelets actively wind or loop fibrin fibers. The mathematical model, while complementary and well-constructed, relies on multiple assumptions and lacks predictive validation.

Appraisal:

While the authors successfully document intriguing fibrin architectures and provide a compelling descriptive framework, they do not fully demonstrate a mechanistic model of active fibrin winding by platelets. The conclusions regarding platelet-driven winding and rotational dynamics are not sufficiently supported by direct or quantitative evidence. To substantiate these claims, the study would benefit from experiments that directly link platelet dynamics to fibrin organisation, including coordinated measurements of platelet motion and fibre rearrangement. As it stands, the results are suggestive but do not definitively support the proposed mechanism.

Discussion and Impact:

Despite these limitations, the study addresses an important question in thrombosis and hemostasis and introduces a potentially impactful conceptual framework for understanding clot compaction. The imaging approaches and datasets presented will be valuable to the community, particularly for researchers interested in platelet mechanics and fibrin organisation. However, the overall impact will depend on whether the proposed mechanism can be more rigorously validated. In its current form, the study presents an interesting and thought-provoking model, but would benefit from either stronger experimental support for the proposed mechanisms or a more cautious interpretation of the findings.

Reviewer #3 (Public review):

Summary:

This work aims to understand the mechanisms that platelets use to interact with and compact fibrin fibers during clot formation. This is an important process during wound healing, and recent work has demonstrated that platelets play a critical role in generating the force required to drive the accumulation of fibrin. The authors argue that current models are insufficient to account for the observed reduction in clot volume and propose that platelets actively 'wind up' these fibers by undergoing myosin-dependent rotation. While interesting, the experiments performed by the authors do not directly test this mechanism, and further evidence is required to support their claims.

Weaknesses:

(1) The motivation to switch from the system used in Figures 1 and 2 to the '2D fiber-retraction assay' is not clear. While the authors state that this system has 'reduced complexity', the differences between these assays appear to disrupt the 'cage-like' organization of fibrin around platelets shown in Figures 1 and 2 (compare images in Figure 2 with those in Figure 4). An in-depth comparison of two methods is needed to support the conclusions from the 2D system. Furthermore, the change in plasma volume (Figure 2 vs Figure 7) should also be tested - the authors state that this increases fibrin fiber formation, but this is not quantified or demonstrated in the figures. Notably, this appears to change the morphology of the fibrin fibers shown (comparing Figure 2 and Figure 7).

(2) It is unclear how the classification of platelets as 'fiber-winding' versus 'fiber compaction' differs in Figure 2. The criteria used for these classifications should be stated. Further, it seems premature to characterize fibers as wound without having established this earlier in the manuscript.

(3) Is the 'gearwheel' different from the 'cage' of fibrin fibers? They appear similar, but it is difficult to distinguish between them with only qualitative descriptions of these phenotypes.

(4) The quantification of platelet extensions in Figure 9 is confusing. While those in 9A are clear, those in 9B are not. For instance, what is the difference between #7 and #8 in the middle panel of 9B? It does not seem like #8 is labeling an extension.

(5) It is unclear what the modeling accomplishes, as there is no comparison between the results of these simulations and their experiments.

(6) The data presented in Figure 12 provides the most direct support for their mechanism, but falls short of directly testing their claims. These experiments should be repeated to include blebbistatin to test the contribution of myosin and include quantitative rather than qualitative comparisons of these experiments.

Author response:

The following is the authors’ response to the original reviews.

We thank the editor and the reviewers for their time and efforts to evaluate our manuscript. We have taken into account all the comments and revised the manuscript accordingly which has considerably strengthened the message.

Several parts of the manuscript, have been extensively rewritten to add explanations and clarify our hypothesis and claims, this also led us to add four new references.

In addition, we have added the following new figures:

- New part of figure 2. Figure 2E shows a platelet in a constrained clot after 4h of retraction with the fibrin cage around the platelet center still present. The actin staining of the platelet shows that radial actin fibers are present in each bulb extending to the platelet center. This observation supports our hypothesis that in each bulb an individual cytoskeletal swirling could take place resulting in the accumulation of fibrin fibers at the base of each bulb.

- Modification of figure 3, to include the criteria used to define four categories of platelets and associated fibers in the 2D fiber retraction assay (new Fig. 3C).

- New figure 13, illustrating the quantification of fibrin fibre compaction mediated by platelets in the 2D fiber retraction assay and the rotational movement of a fiber mass (video 9).

- New supplementary figure 1, showing the result of a new model simulation in the absence of cytoskeleton swirling. Under this condition the fibrin fiber does not loop around the platelet bulb.

Public Reviews:

Reviewer #1 (Public review):

This paper reports a previously unrecognized mechanism by which platelets compact fibrin fibers during clot retraction. Rather than simply pulling on fibers, the authors propose that platelets generate swirling motions that wind and loop fibrin into dense structures.

While the results are intriguing, the underlying physical mechanism remains unexplained. In particular, it is unclear how platelets generate swirling motion capable of inducing fibrin coiling, especially when suspended in 3d fibrin mesh. This raises concerns about the conclusions.

The reviewer is right, it is difficult to imagine how platelets in a 3D fibrin mesh can accumulate fibers at the base of their extensions to form a cage-like fiber organisation around the center of the platelets. We therefore developed the 2D fibre-retraction assay, which we believe provides important insight for the coiled fiber accumulations above spread platelets in the 2D situation but also provides a framework for interpreting similar processes that may occur within a 3D clot. In response, we have placed greater emphasis on clarifying and strengthening the comparison between the potential mechanistic aspects in the 2D and 3D assays, in order to better support our proposed model (see Results, section: "Platelets, spread on a 2D surface, organize fibers above them", last paragraph). In addition, the Ideas and Speculations section of the discussion has been extensively rewritten to provide more detailed explanations about the potential mechanism leading to fibrin fiber accumulations around platelet bulbs in a 3D fibrin mesh.

Also, does fibrin have inherent chirality or structural asymmetry that could promote coiling independently of platelet activity?

Yes, double-stranded fibrin protofibrils have a helical twist [1]. Furthermore, a clot formed in the absence of platelets and other cellular components shows intrinsic tensile forces [2]. However, we show that inhibition of actomyosin actions prevents fibrin fiber accumulation in the 2D fibre-retraction assay providing evidence that platelet actions are necessary to observe the coiled fibers above spread platelets. This has been accentuated in the revised version and three references have been added.

Furthermore, platelet retraction typically involves platelet aggregation rather than isolated cells, and it is unclear how fibrin coiling would proceed in clustered platelets.

Under the in vitro fiber retraction conditions used in our study (constrained or unconstrained clots or even in the 2D assay) individual platelets are homogenously distributed within the forming clot or on the coverslip. Therefore, there are no big platelet aggregates or clusters of platelets under our experimental conditions and the results can only demonstrate how individual platelets act on fibrin fibers. This point has been emphasized in the revised version (Discussion, third paragraph).

Reviewer #2 (Public review):

Summary:

Grichine et al. investigate platelet-mediated fibrin compaction using human donor platelets and propose a novel mechanistic model in which platelets generate contractile forces and wind fibrin fibers into compact coiled structures. Using a combination of 2D spread assays, 3D clot imaging via expansion microscopy, live-cell imaging, and computational modelling, the authors present evidence of cage-like fibrin architectures, coiled-fibre morphologies, and platelet centred "rosette" structures present during fibre compaction. They further suggest that actomyosin-driven cytoskeletal dynamics, potentially involving rotational or swirling motion, underlie this proposed winding mechanism, analogous to DNA looping and compaction. The study addresses an important and longstanding question in thrombosis and hemostasis and offers a conceptually novel perspective on clot compaction.

Strengths:

The integration of multiple imaging modalities is a notable strength of this paper. In particular, the 2D fiber-retraction assay provides a useful model for understanding the spatio-temporal dynamics of platelet-mediated fibrin compaction, which can be applied to other systems and may yield detailed mechanistic insights into biological processes. The live-imaging approaches are particularly well executed and offer valuable dynamic insight.

Weaknesses:

The primary weakness of this paper lies in its descriptive nature and its reliance on correlative rather than causal evidence. Several interpretations are not uniquely supported by the data presented. For example, the categorisation of fibrin accumulation in 2D assays as "fiber winding" and "fibre compaction" remains descriptive without establishing winding as a mechanism.

When introducing the 2D fiber-retraction assay (figure 3) in the revised version, we now only mention the terms fiber accumulation and compaction to better align with the level of evidence, since wound-up fibers cannot be distinguished in this figure. The criteria to establish the four categories of platelets and associated fibers in the 2D fiber retraction assay have now been included in figure 3C.

Nevertheless, coiled fibers above spread platelets are clearly visible in figure 4 and 8 and dynamic fiber rotations or winding-up are observed in figure 12 and video 9. These observations have been presented more cautiously, as indicative rather than definitive evidence of a winding mechanism.

Alternative mechanisms, such as circular bundling, stacked fibers under tension, or fibrin crosslinking-induced aggregation, are neither excluded nor investigated.

For fibrin fiber bundling, staggered or crosslinked protofilaments no platelet actions are necessary as described previously [2,3]. Since we observed a clear difference between +/- blebbistatin conditions in the 2D fiber-retraction assay, the fiber compaction we observe depends on platelet actions. Consequently, we consider these alternative mechanisms unlikely based on our data. This has been stated explicitly in the results section and discussion and three references have been added.

Although the authors present compelling live imaging, establishing winding as a dynamic phenotype would require quantitative analyses, such as measuring angular velocities and coiling rates.

We have incorporated quantitative measurements (new figure 13) about platelet mediated fibrin fiber compaction and angular rotation velocities to complement the observations obtained from live imaging. It is important to note, however, that angular velocities and coiling rates are likely influenced by the number of fiber–fiber contacts present at the time coiling occurs. Specifically, an increased number of contacts is expected to elevate tension within the network, thereby modulating the forces generated by platelets and, consequently, affecting both velocity and coiling dynamics.

The use of a second fluorophore-labelled fibrin population could further strengthen evidence for rotational dynamics.

These live videos are quite difficult to acquire because of the following reasons:

- Small platelet size

- Heterogeneity of platelets within the population (10 d half-life, old platelets may not be able to compact fibers efficiently).

- The speed of the process and the time needed to adjust parameters for image acquisition, necessitates an arbitrary choice of the acquisition window and only one acquisition (90 min) per sample preparation is possible.

- Furthermore, the laser-induced illumination can perturb the observed processes. We therefore use high-spatial-resolution 3D confocal time-lapse imaging, performed in photon-counting mode with very low laser excitation.

For these reasons, the use of additional markers would be technically challenging and could perturb the delicate equilibrium and dynamics of the process under investigation.

Similarly, the inference of rotational contractility or actomyosin "swirling", based on chiral actin organisation and blebbistatin treatment, is not sufficiently supported to conclude that platelets actively wind or loop fibrin fibers.

Importantly, in the 2D fiber-retraction assay, we do not propose that the rotational actomyosin activity leads to a contractility of the platelets which would allow fiber retraction. Rather, we suggest that cytoskeletal actomyosin swirling (as demonstrated for nucleated cells by Bershadsky's team) can induce rotational dragging of extracellular bound fibrin fibers around the pseudonucleus of spread platelets thereby promoting accumulation of fibrin fibers (shown in figure 12C, video 9, third panel). Consistent with this interpretation, inhibition of myosin by blebbistatin prevents the accumulation of fibrin fibers above spread platelets in the 2D fibre retraction assay (Fig. 3).

The mathematical model, while complementary and well-constructed, relies on multiple assumptions and lacks predictive validation.

We thank the reviewer for this insightful comment and acknowledge that the proposed model relies on several important assumptions. In our view, the most significant assumption is that integrin molecules undergo rotational downstream motion as a consequence of their coupling to the swirling cytoskeleton. To assess the necessity and impact of this assumption, we provide an additional simulation performed in absence of the cytoskeletal swirling. Under this condition the fibrin fibers are not looped around the platelet bulb (this result has been added as supplementary figure 1). This analyses also provides further validation of the proposed model and underlying mechanism. At the same time, it is important to emphasize that the primary purpose of the model was to examine whether the hypothetical swirling dynamics of the cytoskeleton, together with the associated receptors, could in principle reproduce the experimentally observed fibrin organization.

Appraisal:

While the authors successfully document intriguing fibrin architectures and provide a compelling descriptive framework, they do not fully demonstrate a mechanistic model of active fibrin winding by platelets. The conclusions regarding platelet-driven winding and rotational dynamics are not sufficiently supported by direct or quantitative evidence. To substantiate these claims, the study would benefit from experiments that directly link platelet dynamics to fibrin organisation, including coordinated measurements of platelet motion and fibre rearrangement. As it stands, the results are suggestive but do not definitively support the proposed mechanism.

Discussion and Impact:

Despite these limitations, the study addresses an important question in thrombosis and hemostasis and introduces a potentially impactful conceptual framework for understanding clot compaction. The imaging approaches and datasets presented will be valuable to the community, particularly for researchers interested in platelet mechanics and fibrin organisation. However, the overall impact will depend on whether the proposed mechanism can be more rigorously validated. In its current form, the study presents an interesting and thought-provoking model, but would benefit from either stronger experimental support for the proposed mechanisms or a more cautious interpretation of the findings.

We agree that the proposed mechanism requires further validation. In the revised version we have added a new result (figure 2E) showing that radial actin filaments are present in each bulb of a platelet in a constrained clot, supporting the possibility that rotational cytoskeletal movements could take place in individual bulbs. In a new figure 13, we have also quantified fiber compaction and the angular velocity of a rotating fibrin mass observed in video 9. Furthermore, in the revised manuscript, we present a more cautious and explicitly hypothesis-driven interpretation of the mechanism. We hope that the publication of our observations will be of interest to researchers in the field of thrombosis and clot mechanics who possess the specialized tools and expertise necessary to rigorously evaluate and either substantiate or refute the proposed mechanistic model.

Reviewer #3 (Public review):

Summary:

This work aims to understand the mechanisms that platelets use to interact with and compact fibrin fibers during clot formation. This is an important process during wound healing, and recent work has demonstrated that platelets play a critical role in generating the force required to drive the accumulation of fibrin. The authors argue that current models are insufficient to account for the observed reduction in clot volume and propose that platelets actively 'wind up' these fibers by undergoing myosin-dependent rotation. While interesting, the experiments performed by the authors do not directly test this mechanism, and further evidence is required to support their claims.

We do not "propose that platelets actively 'wind up' these fibers by undergoing myosin-independent rotation" of the whole platelet, but rather of the cytoskeleton winding-up extracellular fibrin fibers attached to integrin receptors.

Weaknesses:

(1) The motivation to switch from the system used in Figures 1 and 2 to the '2D fiber-retraction assay' is not clear. While the authors state that this system has 'reduced complexity', the differences between these assays appear to disrupt the 'cage-like' organization of fibrin around platelets shown in Figures 1 and 2 (compare images in Figure 2 with those in Figure 4). An indepth comparison of two methods is needed to support the conclusions from the 2D system.

We agree that the cage-like fibrin organization around platelets is disrupted in the 2D fibre-retraction assay when platelets are completely spread on the coverslip before they have encountered fibrin fibers (Fig. 4). This has been explicitly stated in the revised version. However, some platelets in the 2D fiber-retraction assay form the same number of extensions as platelets in a 3D clot (Fig. 9 A, B) and are not completely spread on the glass surface. For these platelets a cage-like fibrin organisation is retained under the 2D conditions (Fig. 5 and 6). Nevertheless, the fiber density at the base of the bulbs is higher in the 2D assay than under the constrained 3D clot retraction conditions (Fig. 1C and Fig. 2), probably because in the 2D condition the fibers are less constrained and readily available for compaction.

Furthermore, the change in plasma volume (Figure 2 vs Figure 7) should also be tested - the authors state that this increases fibrin fiber formation, but this is not quantified or demonstrated in the figures. Notably, this appears to change the morphology of the fibrin fibers shown (comparing Figure 2 and Figure 7).

We thank the reviewer for raising this point. We would like to clarify that Figure 2 and Figure 7 correspond to two distinct experimental setups: the constrained clot retraction assay (Figure 2) and the 2D fiber-retraction assay (Figure 7). As such, they are not directly comparable. We understand, however, that the reviewer is likely referring to the apparent differences between Figures 3–6 (lower plasma volume, higher fiber density) and Figures 7–8 (higher plasma volume, lower apparent fiber density).

The reduced number of visible fibers in the latter condition is not solely a consequence of plasma volume per se, but rather results from the formation of a labile fibrin gel at higher plasma concentrations, which is lost during the fixation and aspiration steps. This effect was initially observed across samples from two donors with differing plasma fibrinogen levels. In one case, an unusually low fibrinogen concentration allowed the addition of higher plasma volumes without inducing gel formation. In contrast, in the other sample, a more typical fibrinogen level resulted in gel formation under the same conditions.

Importantly, we performed all experiments using matched donor plasma and platelets. As a result, the precise fibrinogen concentration could not be determined prior to experimentation. Nonetheless, post hoc measurements confirmed that fibrinogen levels in most donor samples fell within the normal physiological range, which allowed us to always use the same plasma volumes for low and high plasma concentrations (4ul/ml PBS and 7 ul/ml PBS, respectively) except for one donor as mentioned above.

(2) It is unclear how the classification of platelets as 'fiber-winding' versus 'fiber compaction' differs in Figure 2. The criteria used for these classifications should be stated. Further, it seems premature to characterize fibers as wound without having established this earlier in the manuscript.

The reviewer probably refers to figure 3 and he is right; it is premature to mention fiber winding at this stage of the results section (see our response to reviewer #2). In the revised version, we have modified figure 3 to include the criteria used to classify the platelets into four different categories (Fig. 3C).

(3) Is the 'gearwheel' different from the 'cage' of fibrin fibers? They appear similar, but it is difficult to distinguish between them with only qualitative descriptions of these phenotypes.

The "gearwheel" is observed for completely spread platelets in the 2D fiber-retraction assay and a figure illustrating our hypothetical speculations to compare the 2D gearwheel with the 3D clot situation is presented in the discussion under the "Ideas and Speculations" paragraph (now Fig. 14). We have given a more comprehensive explanation of the proposed mechanism in the revised version.

(4) The quantification of platelet extensions in Figure 9 is confusing. While those in 9A are clear, those in 9B are not. For instance, what is the difference between #7 and #8 in the middle panel of 9B? It does not seem like #8 is labeling an extension.

For the platelet shown in the middle panel of Figure 9B, the extensions cannot be clearly distinguished in the MIP (Maximum Intensity Projection) image because extension #8 is positioned above extension #7 and is therefore superimposed in the projection. However, the two extensions can be differentiated when examining the 3D image stack (Video 4, upper panel). As indicated in the figure legend, the number of extensions was determined manually by scrolling through the z-stack image sequence. In the revised version, we will also define the abbreviation “MIP” as Maximum Intensity Projection.

(5) It is unclear what the modeling accomplishes, as there is no comparison between the results of these simulations and their experiments.

We thank the reviewer for this valuable concern. We chose not to combine the experimental fibrin organization and the modeling results within the same figure panel, as the resulting image would be too complex and difficult to interpret. We have, however, added a supplementary figure 3 showing the results of a new simulation in the absence of cytoskeletal swirling. Under these conditions no winding of the fibrin fiber around the platelet bulb can be observed. It is also important to emphasize that the comparison between the model and the experimental data was intended to be primarily qualitative rather than quantitative.

(6) The data presented in Figure 12 provides the most direct support for their mechanism, but falls short of directly testing their claims. These experiments should be repeated to include blebbistatin to test the contribution of myosin and include quantitative rather than qualitative comparisons of these experiments.

As mentioned already above, these live videos are quite tricky to acquire because of the following reasons: - small platelet size

- Heterogeneity of platelets within the population (10 d half-life, old platelets may not be able to compact fibers efficiently).

- The speed of the process and the time required to optimize imaging parameters, necessitate the selection of an arbitrary acquisition window. Consequently, only a single acquisition of approximately 90 min can be performed per sample preparation, with no guarantee that relevant platelet-fibrin interactions can be acquired in the acquisition window.

- Furthermore, after blood donation, the first sample is usually ready to be acquired around 3 pm, acquisition time 90 min. At least 10 successful acquisitions per condition would be required to ensure statistical robustness, but maximal 4 can be acquired per donor, because platelet samples start to deteriorate within twelve hours after blood donation.

Taken together, the intrinsic heterogeneity of the platelet population, the low likelihood of capturing informative events, and the limited availability of suitable imaging resources at our institute render a robust and quantitative comparison between conditions with and without blebbistatin extremely challenging, if not impractical, within a reasonable timeframe.

In accordance with the reviewer's request, we have added a new figure 13 to the revised version, presenting quantitative data on the platelet-mediated fibre compactions and the speed of angular fibrin rotations observed in video 9.

Recommendations for the authors:

Reviewer #3 (Recommendations for the authors):

Throughout the manuscript, it is difficult to map the data presented in the figures to the text in the results section. Often, many subpanels are referred to collectively (for example, 'Fig 4 AE and animation, Video 3' on line 150), and the reader is left to piece together how this data fits into the statements in the results section. More guidance from the authors would help to understand the connection between these data and their conclusions.

In the revised version, we have provided clearer explanations to make it easier to understand the conclusions drawn from the data. Concerning the indication "Fig 4 A-E and animation, Video 3" just means that platelets shown in panels A-E of figure 4 can also be visualized in the animation video 3. We have also put an effort to clearly indicate which figure part is presented in the associated video.

There are also many figures that contain redundant information. The authors should consider revising these figures and including some of these repeated images as supplemental figures.

As noted by the reviewers, our study provides predominantly qualitative observations essentially because it is not obvious to choose parameters which would be pertinent and could be quantified accurately using expansion microscopy. A quantitative analysis would allow to show the quantification and a representative image to describe the phenotypes of platelet-mediated fibre organisations. Without a quantitative analysis, we consider it more appropriate to provide multiple examples, enabling the reader to assess the consistency as well as the variability across repeated observations.

Additional References

(1) Jansen KA, Zhmurov A, Vos BE, et al. Molecular packing structure of fibrin fibers resolved by X-ray scattering and molecular modeling. Soft Matter. 2020;16(35):8272-8283.

(2) Spiewak R, Gosselin A, Merinov D, et al. Biomechanical origins of inherent tension in fibrin networks. J Mech Behav Biomed Mater. 2022;133:105328.

(3) Ramanujam RK, Lavi Y, Poole LG, Bassani JL, Tutwiler V. Understanding blood clot mechanical stability: the role of factor XIIIa-mediated fibrin crosslinking in rupture resistance. Res Pract Thromb Haemost. 2025;9(4):102871.

(4) Gaertner F, Ahmad Z, Rosenberger G, et al. Migrating Platelets Are Mechano-scavengers that Collect and Bundle Bacteria. Cell. 2017;171(6):1368-1382 e1323.

Version 2

Abstract

This study reveals a previously unrecognized mechanism by which platelets retract and compact fibrin fibers. Using a newly developed 2D fiber-retraction assay, we observed an initial “gearwheel” pattern of actin–myosin organization in spread platelets with associated, extracellular fibrin patches that appear to form an initiation complex for fibrin fiber attachment and rearrangement. The final outcome of this process results in two morphologically different platelet populations. Spread platelets with coiled fibers around and above their pseudo-nucleus. Other platelets are only partially spread on the glass surface and are surrounded by tightly packed fibrin fibers around bulbous protrusions (“bulbs”), mirroring the architecture of platelets and adjacent fibers within a retracted clot. Thus, the observed compaction process might also take place during clot retraction in order to reduce clot volume, stiffen the clot and enhance wound repair. Apart from pulling on fibers like on a rope, platelets actively wrap fibrin fibers into compact structures, similar to balls of wool. Besides DNA packaging, this represents a new example of a natural fiber compaction mechanism. Using a combination of 3D clot-retraction and 2D fiber-retraction assays, expansion and electron microscopy, live imaging and mathematical modeling, we show that platelets use an actomyosin-driven motion to gather and loop fibrin fibers around the base of bulbous protrusions. These bulbs form when a platelet becomes trapped between fibrin fibers during 3D clot retraction or 2D fiber-retraction assays. These findings complement and extend earlier models of platelet-mediated fibrin fiber retractions, offering new insight into how platelets mechanically organize fibrin fibers.

Introduction

Platelets are absolutely essential in case of vessel injury. They are not only necessary to initiate hemostasis but also to retract the forming clot thereby pulling together the edges of the ruptured vessel and flattening the clot for improved blood flow. Different approaches to visualize fibrin fibers in whole blood or plasma clots have shown that fibrin fibers represent the main mechanical and structural framework of clots,1,2 occupying a substantial portion of the clot volume. Previous studies have shown that platelets are able to retract an unconstrained plasma clot to a tiny volume.3-6 The extent and rate of clot retraction are influenced by several factors, including platelet and red blood cell counts, fibrin and thrombin concentrations, and the geometric context of clot formation.7 However, platelets are the primary cellular agents driving such extensive clot compaction, although erythrocytes also contribute.8 How do platelets accomplish this tremendous task? Two key studies offer insight: While Fletcher’s lab demonstrated that platelets generate contractile forces comparable to muscle cells,9 Weisel’s group provided mechanistic insight, showing that platelets extend filopodia to bind and pull on fibrin fibers, much like tugging on ropes10 causing localized fiber accumulation.

However, pulling alone may not fully account for the dramatic reduction in clot volume. We hypothesized that efficient clot retraction requires substantial densification of the fibrin fiber network. Using expansion microscopy, we observed that fibrin fibers encircle the central region of platelets within a constrained clot, forming a cage-like structure. Using a “2D fiber-retraction assay” we found that fibrin patches are initially arranged in a ring-like pattern in close proximity to the circular organization of the actin cytoskeleton in spread platelets. Over time, a myosin-dependent mechanism drives the accumulation of coiled fibrin fibers above the plasma membrane of fully spread platelets and their compaction around platelet bulbs in partially adherent platelets.

To explore the mechanistic aspects of the platelet mediated fiber compaction we used computational modeling and live-cell video microscopy. Our results suggest that actomyosin-driven swirling within spread platelets on a 2D surface or within individual bulbs of platelets in the 3D clot space may underlie the observed fibrin coiling and compaction.

Results

Platelets become labeled by fluorescent fibrin fibers during clot retraction

We first estimated the capacity of platelets to retract diluted plasma clots (12% plasma) under unconstrained conditions (Fig. 1A). Under these conditions, 4×107 platelets are able to retract the clot volume to approximately 1.5% of its initial volume (400 μl) within 60 min. Even platelets kept overnight at room temperature before the retraction assay were still retracting the clot to the same extent, however after a longer lag phase. We then used live imaging to follow the reorganization of fibrin fibers during unconstrained clot retraction. Platelet rich plasma (PRP) was spiked with fibrinogen-Alexa 488 before thrombin induced clot formation and image acquisition. Very rapidly one can observe the formation of fluorescent fibrin nodes during the retraction process. Since under the assay conditions the clot is composed only of platelets and fibrin fibers, we hypothesized that the nodes may correspond to platelets surrounded by fibrin fibers (Fig. 1B and video 1). To test this hypothesis, we performed a time course analysis of clot retraction in presence of fluorescent fibrinogen under constrained conditions preventing complete retraction. This approach allowed us to stain and visualize individual platelets in a clot even after prolonged retraction times. Hemostatic clots that form in vivo following vessel injury are also likely to experience mechanical constraints, as they remain anchored to the damaged vessel walls while spanning the breach. Retracted clots were fixed and stained for the αIIb integrin subunit to detect the localization of platelets within the clot. At the position of each platelet, we consistently see colocalization of a fibrin node at all time points examined (Fig. 1C), confirming our hypothesis that the fibrin nodes observed in figure 1B form at platelet sites. Furthermore, the results show that the build-up of fibrin fibers around individual platelets occurs very rapidly, within 10 min. The fibrin fibers forming the nodes did not decorate the entire platelet surface but surrounded the center of platelets embedded in the clot. Please note, that platelets occupy a larger space after longer retraction times, but the fiber nodes persist even after 80 min of retraction.

Platelets become labelled by fluorescent fibrin fibers during clot retraction

A)Time course of unconstrained clot retraction by platelets (4×107 in 400 μl of 12% plasma in PBS) for a total of 60 minutes at room temperature. Experiments have been performed 6x using blood from different donors and typical examples of blood from two different donors are shown (experiments III and IV are repetitions of I and II the day after; although a longer lag-phase is observed, the final retraction volume is similar for the four experiments). B)Unconstrained clot retraction (1×108 platelets per ml in 50% plasma/50% PBS) in presence of fibrinogen-Alexa 488. Image acquisition was started immediately after thrombin addition at a focal plane 100 μm above the bottom of the well and image stacks were collected (61 focal planes, step size 0.5 μm) for a time period of 20:27 min (100 frames with a time interval of 12.5 sec). See also associated video 1. The experiment has been performed twice using platelets of the same donor. Shown are the first, intermediate and last time points of a depth color-coded time-lapse video (left panel; scale bar 10 μm) and the maximum intensity projections (MIPs) of the same time points (right panel; scale bar 10 μm). C)Constrained clot retraction (1×108 platelets per ml in 50% plasma/50% PBS, fibrinogen-Alexa 488 final concentration 12.5 μg/ml) between two holders. Clots were induced by addition of thrombin (2.5U/ml final), fixed at the indicated retraction times, embedded in gelatin, flash frozen and cryosections (14 μm) were stained for the σIIb integrin subunit (magenta; scale bar 5 μm). The time course was performed twice using PRP from two different donors (retraction assays were repeated, although not for all time points, more than eight times using blood from different donors, with consistent results). Image acquisition was performed using a wide-field epi fluorescence microscope (BX41; Olympus) equipped with a Plan 100x/1.25 NA oil objective, a camera (DP70; Olympus), and the acquisition software analySIS (Olympus).

To gain a better insight about the organization of fibrin fibers around platelets within a clot, we used 4x isotropic expansion microscopy. We found that fibrin fibers form a cage-like structure around the platelet center, clearly visible when scrolling through the image stacks (Fig. 2 and video 2 with compiled animations of image parts 2A-E). This pattern was consistently observed across all platelets in the clot and after different clot retraction times (15 min Fig. 2A-D and 4h Fig. 2E). Most platelets had formed large bulbs, thin filopodia, or both with fibrin fibers typically localized around the base of the bulbs or along filopodia. It remains unclear whether all fibrin segments reside on the external platelet membrane or whether some of them extend into channels of the open canalicular system (OCS). Some of these channels persist in activated platelets.11 Please note that straight fibers radiate from the platelet, indicating a high tension in the platelet clot environment. We want to emphasize that changes in temperature, or plasma/platelet concentrations change the kinetics of retraction but we always observe the cage-like fibrin fiber organization around individual platelets (not shown). We also investigated the organization of the actin cytoskeleton of platelets in a constrained clot and demonstrated the presence of radial actin filaments in each bulb, extending to the center of the platelet, where the cage like fibrin fiber organization is observed (Fig. 2E and video 2).

Fibrin fibers are organized in a “cage-like” fashion around platelets within a constrained clot

A-D: Two typical examples (out of n=39 acquisitions of four experiments using blood from different donors) of platelets in a constrained clot with attached fibrin fibers are shown (1×107 platelets per ml PBS/50% plasma and fibrinogen-Alexa 488). Clot retraction was allowed to take place in an inoculation loop (see methods section) for 15 min before fixation and immunofluorescence staining of the σIIb integrin subunit (magenta). Samples were then processed for expansion (scale bars 10 μm = 2.5 μm after correction for expansion; indicated z-levels are not corrected for expansion; see also associated video 2 with compiled animations of image parts A-E). A) Four focal planes from a stack of images (z-level as indicated) showing a platelet in a clot (fibrin fibers in green, plasma membrane staining using an antibody against the σIIb integrin subunit in magenta and the merge). B) D reconstruction of the image stack shown in A (46 planes, step size 0.5 μm), a bulb is indicated by an arrow. C)Four focal planes show another typical platelet in a clot (z-level as indicated). D) 3D image reconstruction of the image stack shown in C (28 planes, step size 1 μm), a bulb is indicated by an arrow. E)3D image reconstruction of a platelet in a constrained clot (retraction in an inoculation loop for 4h) stained after expansion using the NHS-ester (cyan, for total protein stain) and an actin probe (magenta), demonstrating that radial actin fibers extending to the platelet center are present in each bulb (scale bar 10 μm = 2.5 μm after correction for expansion, see also video 2).

Platelets, spread on a 2D surface, organize fibrin fibers above them

So far, our results demonstrate that platelets possess a remarkable capacity to precisely organize fibrin fibers around themselves when embedded in a clot. It is, however, difficult to find out how platelets integrated in the fiber network manage to do that. To investigate the mechanism by which platelets arrange the fibers around their center we sought to use a model system with reduced complexity. We, therefore, developed a “2D fiber-retraction assay”, which is a reductionist approach that does not replicate full clot retraction, but can provide insight into how platelets interact with and organize fibrin fibers. To this end, a diluted suspension of PRP (2.5×106 platelets/ml, 4μl plasma/ml PBS) was incubated with fluorescent fibrinogen and thrombin for 15 min to induce platelet activation and fibrin polymerization. Platelets were then allowed to spread on a 2D surface for 30 min before fixation. Under these conditions, we observe four categories of spread platelets and associated fibrin fiber organizations (Fig. 3; for the criteria used to define the categories see Fig. 3C). The first category comprises platelets without fibrin fiber contact (28.3+2%). The second includes platelets with either a small fibrin dot at their center, which may represent a fibrin initiation complex, or a few fiber contacts above them (40+3%). A third category consists of platelets displaying an initiation complex with attached fibrin fibers and evidence of fiber buildup above the platelets (fiber accumulation 18.3+0.5%). A fourth category exhibits a high fiber densification above platelets (fiber compaction 13.6+0.9%) (Fig. 3A, left image; quantified in Fig. 3B). We then investigated, whether this fiber accumulation/compaction depends on myosin actions as it is the case for clot retraction. Myosin inhibition using blebbistatin strongly diminished the fiber buildup and densification above platelets (fiber accumulation 6.6+0.9% and fiber compaction 0.2+0.2%), while a higher percentage of platelets with fiber initiations or fiber contacts was observed (63.4+2%) (Fig. 3A, right image; quantified in Fig. 3B). For fibrin fiber bundling, staggered or crosslinked fibrin protofilaments no platelet actions are necessary as described previously.12,13 Since we see a clear difference between +/− blebbistatin conditions, demonstrating that platelet actions are required, we consider these alternative mechanisms to be unlikely.

Spread platelets accumulate fibrin fibers above them in a myosin dependent way

A)Shown is a 2D fiber-retraction assay in the presence of DMSO (control) or 50 μM blebbistatin for myosin inhibition. After a 30 min preincubation of platelets (4 μl plasma per ml PBS, see methods section) with or without blebbistatin, thrombin is added and the assay is continued as described in the methods section. Scale bar 10 μm, arrowheads indicate platelets without fiber contact, thin arrows show platelets with a fiber initiation complex or with some fiber contacts. Larger arrows point to platelets with fiber accumulations and even larger arrows to platelets with compacted fibers (see legend below bar graph in B and the detailed features used to classify the different categories described in part C). B)Quantification of the percentage of platelets in the different categories using image acquisitions of four experiments using blood from different donors (for control conditions a total of 636 platelets were counted and for blebbistatin conditions a total of 524 platelets). Shown are means + sem, evaluation of significance using the Student’s t-test. C)Criteria used to classify the platelets and associated fibers shown in A. The green fluorescent fiber images shown in A were used to measure the mean fluorescence intensity of fibers present within a region of interest (ROI; diameter = 5.5 µm) at platelet sites. The measured values are shown below each platelet. The intensity values, together with the associated features (listed on the right), were used to define the four observed platelet categories.

Then, we used the 2D fiber-retraction assay followed by expansion microscopy to investigate in more detail how the fibers around platelets are organized. Platelets were stained either for the αIIb integrin subunit or for myosin. We observed platelets fully spread on the glass surface with coiled fibrin fibers above them (category “fiber accumulation” in fig. 3A). These fibers are frequently wrapped around a bulbous protrusion in the middle of the spread platelet, often referred to as pseudo-nucleus in the literature (Fig. 4A-E and video 3 with the compiled animations of image parts 4A-E). In this situation the cage-like fiber organization is disrupted.

Coiled-up fibrin fibers above spread platelets

Two representative examples (out of n=18 acquisitions from three independent experiments using blood from different donors) of fibrin fiber accumulations above spread platelets in the 2D fiber-retraction assay (4 μl plasma per ml PBS, see methods section) are shown (see also video 3 with compiled animations of image parts A-D). Samples were stained for the αIIb integrin subunit and expanded (scale bars 10 μm = 2.5 μm after correction for expansion, indicated z-levels are not corrected for expansion). A) Three different focal planes (z-levels as indicated) from an image stack showing two spread platelets and attached fibrin fibers. B)3D reconstruction of the image stack shown in A (51 planes, step size 0.25 μm), represented are three different view angles. C)Three different focal planes from an image stack illustrating a spread platelet and attached, coiled-up fibrin fibers. D)3D reconstruction of the image stack shown in C (50 planes, step size 0.25 μm), represented are three different view angles. The fibrin fibers are additionally displayed as depth-color coded (Fire LUT) reconstructions to highlight the straight fiber parts along the z-axis. E)Same fiber image stack as in C and D. Left panels are maximum intensity projections of the green fibrin fibers along the z, x and y axes. Right panels are the same projections color coded in the HSB space (Hue, Saturation, Brightness) according to the orientation of individual fiber segments. The fibers are well aligned along the z-axis, while they crisscross horizontally.

However, in addition to these spread platelets with coiled fibrin fibers above them, we also observed platelets which form bulbs and filopodia similar to platelets within a clot. These platelets were only partially attached to the glass surface and a substantial fiber accumulation and densification was observed around the platelet center at the base of the bulbs (category “fiber compaction” in fig. 3). This fiber organization appeared similar to the cage-like structure formed by fibers around the center of platelets embedded in a clot. Nevertheless, under these 2D conditions, more densely compacted masses of fibers are observed around each bulb than under the 3D conditions of a clot. It is possible that several layers of fibers are tightly compacted around the base of the platelet bulbs, much like individual balls of wool. 6-8 small “balls” can be distinguished in Fig. 5A,C on individual focal planes (see also video 4 with the compiled animations of image parts 5A-D). Additional examples are shown in Fig. 6 (and video 5 with compiled animations of image parts 6A,C,D). The enhanced fibrin labeling observed around these platelets, compared to those in a typical constrained clot (Fig. 2), may be due to the compaction of free, unconstrained fibrin fibers present in the 2D assay. This may suggest that fibers are compacted in multiple loops around platelet bulbs during the progression of their formation. However, the current imaging resolution is insufficient to fully resolve the architecture of these densely compacted fibrin structures.

Strongly compacted fibrin fibers around the center of platelets in the 2D fiber-retraction assay

Two representative examples of platelets are shown with bulbs encircled by fibrin fibers in the 2D fiber-retraction assay (4 μl plasma/ml, see methods section) behaving similar to platelets in a clot (see also video 4 with compiled animations of image parts A-D). Samples were stained for the αIIb integrin subunit (A, B) or the myosin light chain (MLC; C,D) and processed for expansion (scale bars 10 μm = 2.5 μm after correction for expansion, indicated z-levels are not corrected for expansion). Experiment was repeated 4x with blood from different donors with a total of 18 image stacks acquired for fibrin/integrin and 62 for fibrin/myosin staining. A)Three different focal planes from an image stack showing two platelets, one spread and one with bulbs and attached, compacted fibrin fibers. B)3D image reconstruction of the image stack shown in A (34 planes, step size 0.25 μm), represented are three different view angles. C)Three different focal planes from an image stack illustrating another platelet with bulbs and attached, rolled-up, compacted fibrin fibers. D)3D image reconstruction of the image stack shown in C (55 planes, step size 0.25 μm), represented are three different view angles.

Fibrin fibers are twisted in the proximity of platelets

Two representative examples of platelets with compacted fibrin fibers are shown in the 2D fiber-retraction assay (4 μl plasma/ml, see also video 5 with compiled animations of image parts A,C,D). Samples were stained for myosin II (A, B) or phosphorylated myosin light chain (p-MLC; C-E) and processed for expansion (scale bars 10 μm = 2.5 μm after correction for expansion, indicated z-levels are not corrected for expansion). Please note also the strong fiber accumulation around the platelets similar to platelets shown in figure 5. A)3D image reconstruction of an image stack (72 planes, step size 0.25 μm) illustrating a platelet with bulbs and wound-up compacted fibrin fibers. B)Three different focal planes (z-level as indicated) from the image stack shown in A, depth color-coded. Rectangles indicate twisted fibrin fibers and a corresponding zoom is shown below. C) Three different focal planes (z-level as indicated) from an image stack illustrating two platelets with bulbs and attached, wound-up, compacted fibrin fibers. Twisted thick fiber bundles are observed between the two platelets. D) 3D image reconstruction of the image stack (91 planes, step size 0.25 μm) shown in C, represented are three different view angles. E) The same focal planes as in C and the top view of the 3D image reconstruction shown in D of the fibrin fibers, depth color-coded to better distinguish the twist of the thick fiber bundles between the two platelets.

We observe local twisting of fibers in the vicinity of platelets. Particularly clear examples are shown in Figure 6B. These observations may suggest that formation of multiple fiber loops around platelet bulbs induces twisting of the fibers, which in turn may lead to their bundling. Fiber twists can be distinguished in most of the presented images and fiber bundles between two platelets can also be twisted (Fig. 6C-E and video 5 with compiled animations of image parts 6A,C,D).

A ring-like organization of fibrin fiber patches can be observed at the center of spread platelets

We next increased the plasma volume (7μl/ml instead of 4μl/ml) in order to enhance the amount of formed fibrin fibers. This modification led to the formation of a fragile fibrin gel, which was partially lost during the fixation or staining steps. As a result, only very few fibers remained attached to the platelets or to the coverslip.

The residual fibrin fluorescence revealed a ring-like arrangement of fibrin patches at the platelet center. This fibrin “rosette” co-localized with a circular actin organization, forming a gearwheel-like pattern with a compacted fibrin mass at the inner edge of the circle. The actin cytoskeleton was either composed of radial and transverse fibers (Fig. 7 A,B and video 6 with compiled animations of image parts 7A-D) or radial fibers and actin nodules (Fig. 7C and video 6) depending on the degree of platelet spreading.14,15 A faint rosette with a compact fibrin mass and an attached fibrin fiber is shown in figure 7D and video 6.

A fibrin rosette is located close to a ring-like actin organization in spread platelets

Four representative examples (out of n=20 acquisitions, experiment repeated 4 times using blood from different donors) of a fibrin rosette associated with spread platelets in the 2D fiber-retraction assay (7 μl plasma per ml PBS, see methods section) are shown (see also video 6 with compiled animations of image parts A-D). Samples were stained for the αIIb integrin subunit as well as for actin and processed for expansion (scale bars 10 μm = 2.5 μm after correction for expansion). A)3D image reconstruction of an image stack (30 planes, step size 0.33 μm) of fibrin fibers (green) and integrin or actin staining (upper and lower panels respectively, magenta). B)Similar example as in A (17 planes, step size 0.33 μm). C) Another example of a fibrin rosette with intercalated actin nodules (34 planes, step size 0.33 μm). D) An example showing a long fibrin fiber associated with the fibrin rosette (35 planes, step size 0.38 μm).

We then investigated the localization of myosin II under these conditions and found that it is distributed throughout the platelet. However, a denser ring-like organization of myosin is present in the center of the spread platelet, closely associated with the fibrin rosette (Fig. 8 and video 7 with compiled animations of image parts 8A-F). Some myosin spots are intercalated with compacted fibrin (Fig.8 A, focal plane at z=7.8 μm above the culture surface), while coiled fibrin fibers are visible at the periphery and above the myosin ring structure (Fig. 8A-F). The intensity of fibrin rosettes is highly variable between platelets of different donors. Intense rosettes, as shown in figure 7, were seen for platelets from two donors (approximately 3-5% of spread platelets with rosettes), while faint rosettes (Fig. 8AB) were observed for 50% of spread platelets using blood from three other donors.

The fibrin rosette and attached fibers are in close proximity to myosin II in spread platelets

Three representative examples (out of n=62 acquisitions, experiment repeated 4 times using blood from different donors) of fibrin and myosin localization in spread platelets (see also video 7 with compiled animations of image parts A-F) in the 2D fiber-retraction assay (4 μl plasma/ml for A, B, E, F or 7 μl plasma/ml C, D; see methods section). Samples were stained for myosin (A, B, C, D) or phosphorylated myosin light chain (p-MLC; E, F) and expanded (scale bars 10 μm = 2.5 μm after correction for expansion, indicated z-levels are not corrected for expansion). A)Three different focal planes (z-levels as indicated) from an image stack showing a spread platelet and an attached fibrin fiber in the center of a faint fibrin rosette. B) 3D image reconstruction of the image stack shown in A (36 planes, step size 0.3 μm), represented are three different view angles. C) Three different focal planes from an image stack illustrating a spread platelet with a fibrin rosette and an attached, rolled-up fibrin fiber. D) 3D image reconstruction of the image stack shown in C (31 planes, step size 0.33 μm), represented are three different view angles. E) Three different focal planes from an image stack illustrating a spread platelet with attached, rolled-up fibrin fibers. F) 3D image reconstruction of the image stack shown in E (54 planes, step size 0.25 μm), represented are three different view angles.

Modeling platelet-mediated winding-up of fibrin fibers

To address the mechanical aspects of the observed platelet-mediated fibrin fiber organizations, we developed a model based on our experimental observations, which suggests that platelets may be capable of winding fibrin fibers around their bulbs. We first quantified the number of platelet extensions. Each platelet in a clot showed 7-10 bulbs/filopodia (Fig. 9A, C). Similarly, platelets in the 2D fiber-retraction assay, which have formed bulbs and filopodia and show an accumulation of fibers around the base of their bulbs have also 7-10 bulbs/filopodia (Fig. 9B, C). Fiber curvature was determined using transmission electron microscopy (TEM) and fluorescent (FLUO) images. Curvatures were similar for both image types (mean values for κTEM=10.09 ± 6.6 μm-1 and κFLUO=6.77 ± 4.7μm-1) with differences likely due to the higher resolution of TEM (Fig. 9D-F).

Determination of platelet protrusions and fibrin fiber curvatures in proximity of platelets

A)Three examples of platelets within a retracted clot stained for the aIIb integrin subunit (including platelets shown in figure 2). Maximum intensity projections (MIPs) are shown, the number of platelet extensions is indicated and has been determined manually by scrolling through the image z-stack. Scale bars 10 μm = 2.5 μm after correction for expansion. B)Three examples of platelets after 2D-fiber retraction (including platelets shown in figure 5) and stained for the MLC (upper panel) or the aIIb integrin subunit (middle and lower panel). MIPs are shown, the number of platelet extensions is indicated and has been determined manually by scrolling through the image z-stack. Scale bars 10 μm = 2.5 μm after correction for expansion. C)The comparison of all data points for platelets after 3D clot retraction or after 2D fiber-retraction does not show a statistically significant difference between protrusions of platelets in a clot or attached to a glass surface (Mann-Whitney U test). D-F: The degree of fiber curvature in examples shown in D and E is determined as κ=1/r using the radius of a circle fitting the region of strongest curvature of the fibers. D)Three examples of TEM images used to determine the curvature of fibrin fibers close to platelets within a clot (measured examples are indicated by white arrows). E)Three examples of fluorescent images of fibers wound around platelets attached to a glass surface (including platelets shown in figure 5). A single focal plane of each image stack is shown (scale bars 10 μm = 2.5 μm after correction for expansion; measured examples are indicated by white arrows). F) Comparison of all data points obtained using TEM images and fluorescent images (curvature in κ μm-1). For quantification, 31 EM acquisitions of three clots from different donors were used as well as 12 Fluo acquisitions of platelets from 4 different donors.

Our model incorporates these observations, published parameters (Table 1, methods section) and builds on the framework established by Bershadsky and colleagues.16 Bershadsky’s team demonstrated that in fibroblasts, cultured on circular substrates, the actin cytoskeleton is composed of radial actin fibers extending towards the cell center and circular transverse fibers enriched in myosin-IIA. Individual platelets spread on a glass surface have a similar phenotype and cytoskeletal organization. Live-cell imaging in the study by Bershadsky et al. revealed that the transverse fibers slide centripetally along the radial fibers in a myosin dependent process, generating a chiral swirling motion directed towards the cell center. They also show that fibronectin-coated beads bound to integrins on the cell surface are dragged along with the intracellular swirling of the cytoskeleton. Thus, fibrin fibers bound to integrins at the platelet plasma membrane may experience a similar rotational motion around the pseudo-nucleus. Chiral rotations of the cytoskeleton are not only observed at the level of entire cells, but also in subcellular regions, such as individual filopodia of the neuronal growth cone.17,18 Similarly, chiral rotations of the cytoskeleton may be present in individual bulbs of platelets in a clot.

Along this line, we aimed to establish a model which could reproduce the experimentally observed fibrin organization around the base of platelet bulbs due to the hypothetical swirling dynamics of the cytoskeleton and the associated receptors. Our simulations are based on fibrin fibers anchored to integrins in the plasma membrane of platelet bulbs and show that in principle cytoskeletal motions within each bulb can drive integrins, and thus the attached fibrin fibers, towards the platelet center, wrapping the fibers around the bulbs (Fig. 10A and video 8). The efficiency of loop formation depends on the initial fiber position (Fig. 10B-D). The process effectively winds long fibers into compact loops with a radius of ~420-nm around the bulbs (Fig. 10D) and the loop radius is determined by the radius of the bulb (equal to 400 nm in the model).

Simulation of platelet mediated winding-up of fibrin fibers

A) Time evolution of the fibrin fiber configuration around the platelet bulb. Two different angles between the fiber and the x-axis were used ([1]: α = π/4, lime color; [2]: α = π/2.5, dark green color). Left: initial positions, top view and side view. Center: intermediate position, top view and side view. The fibrin fiber began to form a loop around the bulb. Right: final position, top view and side view. A compact fibrin loop was formed around the bottom of the bulb (see also video 8). B) Time when the first node bound to the bulb and thus the formation of the fibrin loop started. The fibrin winding-up started earlier for a larger angle between fiber and x-axis (and consequently smaller angle between fiber and tangent line at the point of the node 0 binding). The number of nodes that were bound to the bulb at the end of the simulation was slightly higher for larger α. Thus, larger α led to a more compact fibrin fiber organization. C)The distance between nodes and the bottom of bulb along z-axis. The distance was smaller for larger α and the fibrin organization was slightly more compact. D)Comparison of final fiber configurations around the bulb for two different α.

In one set of calculations, we switched off the cytoskeleton swirling motion (from Table 1, rate of swirling = 0). In this case, no fibrin loop around the bulb base was formed (supplementary Fig. 1). Thus, it might be concluded that in the proposed system the cytoskeleton swirling is necessary for the fibrin loop formation.

Transmission electron microscopy (TEM) provides evidence of curved and cross-sectioned fibrin fibers at the base of platelet bulbs

To confront the results obtained by expansion microscopy and model simulations with those using TEM, we analyzed TEM images of platelets and fibrin fibers within a clot. We observed numerous instances of cross-sectioned fibrin fibers with a distance between 200-800 nm (similar range as our model predictions) at the base of bulbs as illustrated in Figure 11. This figure also shows two examples suggesting fibers that curve along bulbs from top to bottom and fibers running along a long, thin filopodium.

Features predicted by the model/simulation are observed on TEM images

Interpreted, potential fiber organizations are highlighted in green on an identical image on the right. A)Fibrin fibers appear to encircle a bulb. B)Cross sectioned fibrin fibers at the base of a bulb, which potentially curve around the bulb. C)Several cross sectioned fibers at the base of a bulb, which may surround the bulb. D)Several cross sectioned fibers along a platelet extension potentially forming a spiral. E)Cross sectioned fibrin fibers around a forming bulb. F)Cross sectioned fibers at the base of a bulb and two fibers along a filopodial extension.

Live imaging of platelet-mediated reorganization and compaction of fibrin fibers

We performed live-cell video microscopy using platelets from three different donors to observe how platelets orchestrate fibrin fiber organizations (Fig. 12 and video 9 with compiled timelapse acquisitions associated to image parts 12A,B,C,E). In many cases, fibrin fibers were already accumulated above platelets at the start of imaging, with minimal further fiber movements or compactions; likely due to fibers binding to each other, to the dish or to the platelet surface (Fig. 12A). However, dynamic fiber movements, both clockwise and counterclockwise, were observed above two platelets present within the rectangle indicated in figure 12A (zoom images lower panel). An arrow in Fig. 12A marks the formation of a kink in a fibrin fiber. For the second experiment (blood from a different donor), we reduced the incubation temperature to 21°C in order to slow down the platelet and fiber movements. In the rectangle indicated in figure 12B, a fibrin fiber gets progressively curved and eventually ruptures (at time point 00:48). The resulting free end of the fiber then undergoes a clockwise rotation (viewed from below, zoom images lower panel). Two other platelets producing a fiber densification are marked by an arrow in figure 12B. One of these platelets shows a counterclockwise rotation (viewed from below) of the central part of the platelet during the fiber densification process. A third experiment (platelets from a third donor), again performed at room temperature, shows the rotational (counterclockwise, viewed from below) movement of the pseudo-nucleus and alongside the winding-up of fibers around this bulbous protrusion. This apparently generates strong tension that pulls the pseudo-nucleus towards the edge of the spread platelet, where the movement ceases (Fig. 12C, zoom images lower panel). These fibrin fiber movements occurred without visible filopodia-fiber interactions, suggesting that the bound fibers are dragged by intracellular movements of the cytoskeleton. Since still images cannot fully illustrate the fiber movements or the rotation of the platelet center, we invite the reader to consult the accompanying time-lapse videos (video 9 with the compilation of videos associated to Fig. 12A-C,E). After the time-lapse acquisitions of figures 12B and C, samples were fixed and stained for the αIIb integrin subunit to localize the position of platelets on the dish surface (Fig. 12D). The final fibrin fiber organization around platelets in these live videos of the 2D fiber-retraction assays closely resembles the arrangement of fibers around platelets in a constrained clot (Fig. 1C), except that some platelets are completely spread on the surface with no or very few fiber contacts. The platelets, which had formed bulbs and were partially spread on the dish surface, were surrounded by densely packed fiber accumulations reminiscent of the fibrin fiber network with fibrin nodes colocalizing at platelet positions (Fig. 1C).

Live imaging of spread platelets winding-up fluorescent fibrin fibers

A-C: Shown are three experiments using the 2D fiber-retraction assay with blood from different donors. For each experiment, projections of four focal planes are shown corresponding to the fluorescent fibrin fibers (green), transmission and merged images captured at the first and last time points of the timelapse videos. A)Washed platelets were adjusted to 5×106 platelets per 2 ml PBS and 5 μl plasma was added as well as fibrinogen-Alexa 488. Thrombin was added to induce fibrin fiber formation and platelet activation. After 10 min, the suspension was transferred into a petri dish (WPI Fluoro), centrifuged and installed in the microscope incubator at 37°C to start image acquisition (video 9, upper panel). An arrow in the fluorescent images indicates a forming kink in a fibrin fiber (scale bar 10 μm). The rectangle in the merged images indicates a region where rotational fiber movements take place. Zoom images of this region at different time points are shown below. B)The second experiment was performed under conditions as described in A except that the petri dish was kept at room temperature to slow down the platelet mediated fiber reorganizations (scale bar 10 μm). Arrows in the transmission images indicate two platelets that compact fibrin fibers. The rectangle in the merged images indicates a region where a fibrin fiber is coiled by the platelet and finally ruptures causing the broken fiber to rotate (video 9, second panel). Zoom images of this region at different time points are shown below. C)Platelets of a third donor (not washed, i.e., 5 μl PRP with 4×106 platelets) were added to 2 ml PBS as well as fibrinogen-Alexa 488 and the experiment was continued as in B (scale bar 10 μm). The square in the merged images indicates a region where fibrin fibers are getting coiled and compacted around the pseudo-nucleus of the spread platelet (video 9, third panel). Zoom images of this region at different time points are shown below. The last zoomed image shows a temporal color-coded projection of the transmission channel to illustrate the rotational movement of the pseudo-nucleus (scale bar 5 μm). D)Samples shown in B and C were fixed 3h after the start of time-lapse acquisitions and stained for the aIIb integrin subunit. Shown are fibrin fibers (green) integrin staining (magenta) and the merge (samples B and C, upper and lower panels, respectively; scale bar 10 μm). E) Preformed fluorescent fibrin fibers were prepared in the absence of platelets. PRP was adjusted to 2.5×106 platelets per ml with PBS and 1800 μl was transferred into a petri dish (WPI Fluoro). The dish was centrifuged to allow spreading of the platelets and the supernatant was replaced by the preformed fibrin fibers. The dish was installed in the microscope incubator to start image acquisition (video 9, last panel). The experiment was performed twice with blood of the same donor. The last fluorescent image of the time-lapse video is shown (left image). Fifteen of the fibrin accumulations in the middle of the platelets are rotating in a counterclockwise direction ( indicated as “1”), four rotate clockwise (“2”) and for 16 no clearly visible turn is observed (“3”). A projection of all fluorescent time points, temporal color-coded using the LUT spectral, is shown. White corresponds to the sum of all the colors at each time point, meaning no movement during the time-lapse video (right image; scale bars 10 μm).

To minimize pre-imaging fiber buildup, we next added preformed fibrin fibers to pre-spread platelets and started imaging immediately. Under these conditions, the glass surface is coated and only very few fibers are present. However, a central fibrin accumulation is consistently observed on each platelet, potentially serving as an initiation site for subsequent fibrin fiber organization. Among 35 platelets analyzed, 15 displayed a central fibrin mass rotating counterclockwise (viewed from below), 4 rotating clockwise, and 16 showed no consistent directional movement (Fig. 12E). In some instances, the rotating complexes stalled and stopped moving. These observations support the idea that cytoskeletal swirling could drive fibrin reorganization above spread platelets.

The fiber compaction process mediated by three platelets (two in figure 12B, one in figure 12C, indicated by arrows in the transmission images) was quantified and a fibrin accumulation of a fiber length equivalent to 20, 30 and 70 μm (accumulation rate 1, 0.51 and 1.61 μm/min, respectively), was estimated. Accumulation ceased at an approximate total fibrin fiber length of 80-90 μm (Fig. 13 A-C).

Quantification of platelet induced fiber compaction and angular rotation using the time lapse videos shown in figure 12 (video 9)

A-C: Quantification of fibrin accumulations induced by 3 platelets during time-lapse confocal 3D video microscopy. In each row first and last frames of the series are shown as Z-projections of fibrin fluorescence, followed by the transmitted light image of the platelet in the last frame. The integrated fluorescence intensity in the ROI was quantified as the sum of all Z-planes, and normalized to the intensity of a 1 μm long segment of a single fibrin fiber. The equivalent fibrin fiber length accumulation is assessed as a slope of the linear regression of the initial raising part. The ROI diameter is 3 μm. A)Platelet shown in figure 12B upper left side (indicated with an arrow in the transmission image) B)Platelet shown in figure 12B middle (indicated with an arrow in the transmission image) C)Platelet shown in figure 12C (indicated with an arrow in the transmission image) D)Quantification of the angular rotation speed of a fibrin mass attached to a spread platelet during time-lapse confocal 2D video microscopy (region indicated by a rectangle in Fig.12E, left image). The right platelet presents a fibrin mass attached to the platelet surface, which undergoes a counter-clock wise rotation. Three consecutive frames of the series are shown (time interval 3 min, total duration 240 min, elapsed time as indicated), where rotation speed was maximal and the transmitted light images are shown below. The transmission images show two rotating centers present in this platelet and the fibrin mass is turning around the left center which is within the elongated spread part of the platelet indicating that lamellar protrusions may acquire a new subcellular cytoskeletal unit similar to individual bulbs of platelets in a clot. The total number of turns of this fibrin mass was about 3.3, resulting in the average angular speed of 5°/min with short-term bursts up to 18°/min.

The angular rotation speed of a fibrin mass around a cellular turning center was calculated to be 5°/min with short-term bursts of 18°/min (right platelet within the rectangle indicated in figure 12E, left image). The transmission images show two rotating centers present in this platelet and the fibrin mass is turning around the left center which is within the elongated spread part of the platelet. It is therefore possible that in lamellar protrusions a subcellular cytoskeletal unit is formed, which undergoes individual rotational movements. Formation of transient secondary centers can be seen in other platelets of this video and in a study by Gaertner et al. showing a video of platelets migrating on a fibrinogen substrate which they collect during the migration process (see video mmc6 of their publication19).

Discussion

Using a 2D fiber-retraction assay, we identified a previously unrecognized capacity of platelets to compact fibrin fibers into a small volume possibly by winding them into coiled structures during platelet induced fiber retractions. To our knowledge the only example of natural fiber compaction is coiling of DNA molecules.20 In contrast to intracellular, enzyme-mediated DNA compaction, extracellular fibrin fiber compaction requires the cellular activity of platelets. Our investigation began with the observation that platelets become labeled with fluorescent fibrinogen during clot retraction as previously shown by Brzoska et al.21 They demonstrated that fibrin binding to thrombin activated platelets depends on αIIbβ3 integrins and the actin cytoskeleton. While fibrin fiber retraction by platelets in a clot has been attributed to non-muscle myosin II,22,23 the mechanism underlying fiber organization and compaction has remained unclear.

One key mechanism, revealed in previous studies, is the extension of platelet filopodia that bind to fibrin fibers and pull them inwards in a hand-over-hand motion.10 This is clearly the main process by which platelets retract fibrin fibers in a clot. However, this mechanism alone cannot explain the cage-like fiber organization observed around platelets in a constrained clot (Fig. 2). Our findings introduce an additional, distinct mechanism by which platelets can organize fibrin fibers. Platelets compact fibrin fibers into dense, ball-like structures, which is an efficient form of compaction (Fig. 4-6). Thus, our results reveal an additional mechanistic aspect of platelet mediated fiber retraction beyond those described by Michael et al.24 and Kliuchnikov et al.,25 so far observed only in the 2D fiber-retraction assay. Nevertheless, the platelet morphology and the organization of fibers around them after the retraction process closely resemble those of platelets in a clot and both, 2D and 3D fiber retractions, depend on myosin actions. Thus, it is unlikely that fibrin fibers, which present a helical twist and an inherent tension,13,26 could spontaneously form these ball-like structures independently of platelet activities.

The 2D fiber-retraction assay may recapitulate early post-injury events, such as activation of platelets in suspension triggered by agonists released from neighboring platelets, ultimately leading to the spreading of activated platelets onto the subendothelial matrix to anchor the developing clot at the site of vessel injury. This assay revealed two key observations:

First, when many fibers are present, some platelets spread with the typical fried-egg phenotype and coil up fibers above their pseudo-nucleus (Fig. 4). Others form large bulbs in several directions, similar to platelets in a clot, and compacted fibers are observed around the base of each bulb (Fig. 5, 6). Under these unconstrained 2D conditions, fiber accumulation around the bulbs is much higher than in the case of a constrained 3D clot (Fig. 2), where internal tensions limit fiber accumulation around the base of platelet bulbs. It is noteworthy to emphasize that under all assay conditions used (constrained or unconstrained clots, 2D fiber-retractions) only homogenously distributed platelets are investigated. Thus, the results only show how individual platelets and not platelet aggregates act on fibrin fibers.

Second, when fewer fibers are retained on the culture surface, several spread platelets display a fibrin rosette around their pseudo-nucleus (Fig. 7,8) that colocalizes with actin, myosin and integrins. A dense fibrin mass is seen at the inner periphery of this rosette, sometimes with an attached fiber. This gearwheel-like architecture may represent an early priming/nucleation step of platelet-induced winding of fibrin fibers. The actin cytoskeleton in platelets exhibits a chiral organization27 which, combined with the action of myosin, could be responsible for the rotational motion observed in spread platelets (Fig.12 E and video 9). Integrin-bound extracellular fibers are pulled along by this vortex inducing their coiling around the pseudo-nucleus (Fig. 12C). Fibrin rosettes may fade as coiled fibers become more prominent, suggesting a transition from an initiation phase to active coiling and compaction.

Ideas and Speculations

How this 2D “gearwheel” translates to the 3D clot environment remains unclear. One possibility is that the pseudo-nucleus of spread platelets may correspond to a bulb around which fibers are coiled and accumulate. Platelets with bulbs within a clot have been described previously.28-31 We wondered whether platelet bulbs might represent plasma membrane blebs similar to membrane protrusions formed by megakaryocytes during clot retraction.32 However, the consistent presence of 8–9 bulbs per platelet (Fig. 9A-C) shows the three-dimensional morphology of individual platelets embedded in a fibrin clot, reminiscent of the eight corners of a cube, allowing platelets to sense the clot’s 3D environment along the main directions. Unlike blebs, these bulbs often contain organelles such as mitochondria (Fig. 9D, upper image and Fig. 11B) suggesting that each bulb may have its own energy supply and actomyosin organization. This could enable swirling within each bulb and explain the observed fibrin fiber accumulation at their base. In support of this hypothesis, we have shown that radial actin fibers extending to the platelet center are present in each platelet bulb within a constrained clot (Fig. 2E). Thus, at the onset of clot retraction, each emerging bulb may act as a discrete functional unit onto which the gearwheel assembles, enabling fibrin fibers to attach and initiating their subsequent winding and compaction (scheme, Fig. 14). These bulbs are not static structures, but are constantly moving and renewing. If fibrin fibers bind to integrins on the bulb plasma membrane without subsequent transport of these adhesion complexes towards the bulb base and recycling of integrins back to the top, both bulb remodeling and fibrin network compaction would likely be impaired. The rotational cytoskeletal motion that drags attached fibers towards the bulb base cannot be interpreted as a simple winding of individual fibers, since platelets are embedded within a highly interconnected fibrin network. Instead, rotational transport of bound fibers within this network may generate constrictive forces at the bulb base, where fibers progressively accumulate. This accumulation could promote local fiber densification and thereby contribute to compaction of the fibrin network as a whole.

Schematic illustration of the hypothesis explaining the potential mechanism by which platelets may wind-up fibrin fibers and compact them.

Represented is a platelet in the 2D fiber-retraction assay (upper part) or in a clot (lower part). At the beginning of platelet fiber interactions, a fibrin rosette is present around the pseudo-nucleus and in each bulb with a dense fiber mass at the internal periphery of the rosette possibly representing an initiation complex. Radial actin fibers extend to the center of the platelet and an actomyosin rotational movement along the radial actin fibers may lead to the winding-up and the downward sliding of the extracellular attached fibers to encircle the pseudo-nucleus or the base of the bulbs at the final state of compaction (3D rendering of the scheme was produced using an open AI).

Indeed, blebs of megakaryocytes retracting a clot may behave similarly as platelet bulbs since fibrin fibers can be seen surrounding the base of blebs forming a cage-like structure around the center of the megakaryocyte during megakaryocyte-mediated clot retraction, as observed by Kim et al. (see the supplementary video 3: blood_bld-2023-021545-mmc3 of their publication).32

Our live time-lapse videos of platelet-mediated fiber movements in a 2D environment support the concept of a cytoskeletal vortex that drives the coiling and compaction of extracellular attached fibers (Fig. 12A-C). Additionally, the last time-lapse video capturing the rotation of compacted fibrin clusters at the center of platelets (Fig. 12E) further suggests the presence of cytoskeletal swirling within spread platelets. During fibrin fiber formation this swirling may play a role in scanning the extracellular space for forming fibers that can be anchored to the initiation complex. The process occurs on a notably slow time scale, which could facilitate effective fiber capture by the rotating complex. However, this rate appears insufficient to account for the rapid labeling of platelets during clot formation, suggesting that regulatory mechanisms, possibly involving mechanical feedback, may be at play. Indeed, as mechanosensory cellular particles, platelets adjust their contractile forces and dynamics in response to environmental cues such as fiber tension and clot stiffness.9,10,33,34,25

In conclusion, our study points to a novel mechanistic aspect of platelet function: the potential active winding-up and compaction of fibrin fibers. This concept complements the existing model of platelet-mediated clot retraction10 and provides a new perspective on how platelets may contribute to clot architecture, mechanical stability, and wound healing.

Materials and methods

Reagents and antibodies

Reagents: Neutral formalin (Sigma-Aldrich, Saint-Quentin Fallavier, France), Pluronic F-127 (Sigma-Aldrich, P2443), thrombin (Sigma-Aldrich, T4648), Jasp-K4-HA (a HAK-actin probe, kind gift of Paul Guichard/Virginie Hamel),35 apyrase (Sigma-Aldrich, A6535), heparin (Sigma-Aldrich, H3393), PGI2 (Sigma-Aldrich, P6188), NHS-Ester DyLight 650 (Thermo Scientific) Antibodies: mouse anti-integrin αIIb (Santa-Cruz, sc-7310), monoclonal rabbit anti-HA (Cell Signaling, C29F4), rabbit anti-non muscle myosin II heavy chain (Covance, PRB-440P), rabbit anti-phospho S19 myosin light chain (Abcam, ab2480), monoclonal rabbit anti-myosin light chain 2 (Abcam, ab92721), Cy3 goat anti-mouse IgG (Jackson, 115-165-166), Cy3 goat anti-rabbit IgG (Jackson, 111-165-003), Alexa 647 goat anti-rabbit IgG (Jackson, 111-605-003)

Preparation of human platelet rich plasma (PRP) and platelet free plasma (PFP)

Whole blood was obtained from the French blood bank (age-range 18-65, mixed genders; pooled blood from multiple donors has not been used). The IRB has authorized our institute’s research projects using human samples under the number DC-2022-4976. PRP was prepared from whole blood drawn into Na-citrate Vacutainers (3 tubes of 2.7 ml per donor). The blood is pooled into a 15 ml falcon tube and centrifuged for 12 min 400g, RT, no brake. The upper phase is the platelet rich plasma (PRP), in which platelets are distributed in a gradient (low concentration on top, high concentration at the bottom). In order to increase the platelet concentration in the final PRP, the upper 3 ml are removed and used to prepare platelet free plasma (PFP, by centrifugation at 12000g, 1 min, RT). The remaining upper phase is collected as the final PRP and the platelet concentration determined.

Platelet washing procedure

PRP was diluted 1/4 with Tyrode/0.35% BSA buffer and platelet activation inhibitors (0.02 U/ml apyrase, 1 μm PGI2, 10 U/ml Heparin, final concentrations) were added. Samples were then centrifuged for 15 min at 900g at RT. The supernatant was aspirated, and platelets were gently resuspended in fresh Tyrode/BSA buffer, 0.02 U/ml apyrase, 1 μm PGI2. The suspension was centrifuged as before and the platelet pellet was resuspended in fresh Tyrode/BSA buffer without inhibitors.

Clot retraction assays

We used three different clot retraction assays illustrated in supplementary figure 2.

Unconstrained clot retraction: See paragraph “Live video microscopy” below for method details.

Constrained clot retraction between two holders: In order to perform immunofluorescence staining of individual platelets within a clot, we used a clot retraction assay around two sterile inoculation loops as previously described.36,37 Briefly, PRP was adjusted to 1×108 platelets per ml in 50% plasma/50% PBS (to reduce fibrin fiber density), fibrinogen-Alexa 488 final concentration 12.5 μg/ml and 4 μl/ml erythrocytes for color contrast were added. 400 μl of this suspension was used for clot formation by addition of thrombin (2.5U/ml final) and immediately transferred to 2 ml Eppendorf tubes coated with 2% Pluronic® and containing the inoculation loops. After different retraction times, clots were fixed for 1 h with isotonic formalin (9 volumes formalin / 1 volume 10xPBS). Clots were then washed with PBS and incubated ON at 4°C in 1 ml PBS/15% sucrose. Clots were further incubated in 1 ml PBS/15% sucrose/7.5 % gelatin at 37°C for 4 h and then inclusion blocks were formed in the same solution on ice. Blocks were then trimmed and snap frozen for 1 min in isopentane at minus 60°C and stored at minus 80°C until 14 μm thick transversal sections were cut using a cryostat.

Constrained clot retraction within an inoculation loop: Small clots of 30 μl platelet suspension (50% plasma/50% PBS; 1×107 platelets/ml; fibrinogen-Alexa 488 final concentration 12.5 μg/ml; 4 μl/ml erythrocytes for color contrast) were induced to form within plastic inoculation loops (4 mm internal diameter, attached to a glass coverslip) by addition of thrombin to a final concentration of 2.5 U/ml as described previously.15 Five minutes after clot induction, clots were covered with 500 μl PBS and kept in a cell culture incubator for 10 min. Clots were then fixed with 500 μl isotonic formalin for 15 min, RT, in the dark and then kept in PBS at 4°C until further use. Forming a clot within a small inoculation loop, allowed us to do expansion microscopy without performing cryosections of larger clots, which are difficult to detach from slides during the expansion process.

2D fiber-retraction assay

PRP was diluted in PBS (2.5×106 platelets/ml, plasma concentrations were kept at 4 or 7 μl/ml depending on the amount of fibrin fibers to be formed; see explanation below) and fibrinogen-Alexa 488 was added to a final concentration of 12.5 μg/ml. To induce fibrin fiber polymerization and platelet activation, thrombin was added (final concentration of 2.5U/ml). The suspension was gently mixed once (to prevent platelet aggregation) and after 15 min of incubation at RT in the dark, 400 μl was transferred into wells of a 24 well plate containing glass coverslips. The plate was centrifuged 3 min, 600 g, RT to allow synchronized contact of all platelets with the glass surface. The plate was then placed in a cell culture incubator for 30 min to allow platelet spreading. Samples were then fixed with isotonic formalin for 15 min at RT and then kept in PBS at 4°C until further use.

To promote fibrin fiber formation, while preventing the development of a clot, the plasma concentration must be kept low (4 μl plasma per ml PBS). This can be achieved by adding a volume of PRP containing 2.5×106 platelets to 1 ml PBS, provided the platelet concentration of PRP is at least 6.25×108/ml (condition used for Fig. 3, 4, 5, 6, 8ABEF). If the platelet concentration in the PRP is lower, platelets must be washed and the appropriate amount of plasma added separately.

At higher plasma concentrations (7 μl plasma per ml PBS), more fibers polymerize leading to the formation of a fragile clot which is aspirated during fixation/washing procedures. Under these conditions very few fibers are left behind. However, a fibrin rosette can be observed in the center of several spread platelets (condition used for Fig. 7, 8CD).

The 2D fiber-retraction assay followed by immunofluorescence staining and expansion has been repeated 10x with blood from different donors. A total of 18 image stacks stained for fibrin/integrin, 20 stained for fibrin/integrin/actin and 62 stained for fibrin/myosin have been acquired.

Immunofluorescence

Fixed platelets were permeabilized with PBS/0.2% Triton X-100 for 15 min at RT and then incubated with blocking buffer (3% BSA and 10% goat serum in PBS) for 1h at RT. Coverslips were then incubated for 2 h with a primary antibody diluted in blocking buffer, then washed three times and incubated with a secondary antibody diluted in blocking buffer for 2h at RT. After three washing steps, coverslips were either mounted in Mowiol or processed for expansion.

Expansion microscopy

Expansion microscopy was essentially performed according to Gambarotto et al.38 with minor modifications as described previously.37 Briefly, fixed, immunostained platelets or clots on coverslips (12 mm diameter) were incubated in PBS/0.7% formaldehyde/1% acrylamide over night at room temperature. The coverslip was then placed upside down onto an ice cold 35 μl drop of gelation solution (PBS/19% Na-acrylate/10% acrylamide/0.1% bisacrylamide) pipetted onto parafilm on ice immediately after addition of TEMED and ammonium persulfate both to a final concentration of 0.5%. After 5 min of incubation on ice, the samples were transferred to 37°C for 1 h. The coverslip plus gel facing up was then incubated in denaturation buffer (5.7% SDS/0.2M NaCI/50mM Tris, pH9) for 15 min at RT with gentle agitation to detach the gel from the coverslip. The gel was then transferred into an Eppendorf tube with fresh denaturation buffer and boiled for 30 min at 95°C. To expand the gel, it was placed into a large volume of H20 and the water was changed twice before incubation ON at RT in fresh H20 leading to 4x isometric expansion.

Image acquisition

For image acquisition of some of the expanded samples (Fig. 2, 5CD, 6, 7, 8), a confocal microscope ConfoBright (Nikon A1R+MP) equipped with a home-made adaptive optics module was used with a 40x/1.15 NA apo long distance water immersion objective. The geometric optical aberrations were corrected both in excitation and detection light paths in the open loop mode with a deformable mirror (AlpAO DM97-15) inserted between the confocal scanning head and the microscope inverted stand (Nikon Ti2E). Local aberrations were sensed using the metrics of molecular brightness, derived from fluorescence fluctuations, and were iteratively corrected as individual Zernike modes. This configuration allowed to compensate for the refractive index mismatch, the specimen tilt and the eventual light scattering in depth of expanded samples and to bring back the confocal resolution to its diffraction limit in the vicinity of the imaged platelet.

Image acquisition of other expanded samples (Fig. 4, 5AB) was performed with a confocal microscope Dynascope (LSM 710 ConfoCor 3; Zeiss), 63×/1.4 NA Plan Apochromat objective and the acquisition software Zen 2010. The same microscope and software were used for live imaging (Fig. 12) using the Apochromat 40×/1.2 NA water-immersion objective. Axial Z-stacks (step 0.45 μm) were imaged in a time series using the avalanche photodiode detector in photon counting mode at low 488 nm excitation power (0.1% transmission) to limit photobleaching and phototoxicity, the 633 nm laser was set to 0.6% in order to obtain the transmission light images of the platelets. The pinhole was closed to 0.25 Airy units (14 μm) to increase the spatial resolution. The acquisition of transmitted light was concomitant to the fluorescence.

The FIJI software was used for image processing and analysis.39 No treatment that modifies the raw images was applied. FIJI software was used for the Maximum Intensity Projection (MIP) images, 3D reconstructions of image stacks and for the depth-color coded projection image using the rainbow LUT (Fig. 4D), the color code in the HSB space (Fig. 4E) as well as for the depth-color coded images fire LUT (Fig. 6) and the temporal-color coded image spectral LUT (Fig.12E).

Transmission electron microscopy (TEM)

Clots were formed around two holders as described previously.36 Briefly, PRP was adjusted with PBS to 10 % plasma, 1×108 platelets/ml and 4ml erythrocytes per ml were added for color contrast. Clot formation was induced by addition of thrombin (final concentration of 2.5 U/ml). After 1h of retraction at 37°C, clots were fixed using 1,5 ml of 2,5 % glutaraldehyde for 1h at RT. After fixation, a contrast-enhancing step was performed by incubating the samples in a solution containing 1.5% potassium ferrocyanide and 1% osmium tetroxide in 0.1 M sodium cacodylate buffer. Following this, the samples were embedded in EPON resin according to the protocol described by Eckly et al.40 Thin sections (100 nm) were cut and stained with uranyl acetate and lead citrate and examined using a Jeol 1200 Plus transmission electron microscope operating at 120 kV.

Live video microscopy

Unconstrained clot retraction (Fig. 1B):

Thrombin (final concentration of 2.5 U/ml) was added to 0.4 ml of a platelet suspension (50% plasma/50% PBS; 1×108 platelets/ml; fibrinogen-Alexa 488 final concentration 12.5 μg/ml; 4 μl/ml erythrocytes for color contrast) to induce clot formation. After quick and gentle mixing, the suspension was transferred immediately into a well (8-well Lab-Tek, coated 2% Pluronic® to prevent clot attachment to well walls). Acquisition was started immediately at 100 μm above the bottom of the well using the confocal microscope ConfoBright (for microscope details see paragraph image acquisition).

2D fiber-retraction assay (Fig. 12A-D):

Platelets (washed Fig. 12AB or not Fig. 12C) were adjusted to 2.5×106 platelets/ml PBS and 2.5 μl plasma was added and fibrinogen-Alexa488 (final concentration of 12.5 μg/ml). 15 minutes after thrombin induced fibrin fiber formation and platelet activation, 1800 μl of the suspension was transferred into a petri dish (WPI Fluoro Dish, FD35–100). The dish was centrifuged 3 min, 600 g, RT and installed in the microscope incubator at 37°C (Fig. 12A) or room temperature (Fig. 12BCD).

2D preformed fiber-retraction assay (Fig. 12E):

Preformed fluorescent fibrin fibers were prepared in the absence of platelets (2.5 μl plasma/ml PBS; fibrinogen-Alexa 488, final concentration 12.5 μg/ml; thrombin, final concentration 2.5U/ml; incubation 15 min RT in the dark). PRP was adjusted to 2.5×106 platelets/ml with PBS and 1800 μl was transferred into a petri dish (WPI Fluoro Dish, FD35–100). The dish was centrifuged 3 min, 600 g at RT to allow spreading of the platelets. After centrifugation the supernatant was replaced by the preformed fibrin fiber suspension and the dish was installed in the microscope incubator at 37°C.

For all time-lapse videos, image acquisition was started immediately after adjusting the focal position and image parameters (<1 min) using a confocal laser-scanning microscope (ConfoBright or Dynascope, for microscope details see paragraph image acquisition).

The model of fibrin winding-up

The platelet bulb was modelled as a straight cylinder (the radius R = 400 nm, the length = 1500 nm). We assumed that the internal cytoskeleton of the bulb underwent the swirling motion with the angular velocity Rate_swirl as it was shown in16 and simultaneously moved down to the bottom of the bulb with the velocity Rate. Integrins αIIbβ3 on the bulb surface were associated with internal cytoskeletal proteins,41 thus, they underwent the swirling and downstream motion with the same rates.

The fibrin fiber was modelled as it was done previously24,25 with several simplifications. The fiber composed of N equally spaced nodes connected by elastic springs (supplementary Fig. 3A). The spring equilibrium length was L and it was equal to the initial distance between nodes. The fiber was located in the XY-plane and the angle α between fiber and x-axis was variable (Table 1). The motion of the node i in solution was described by the second Newton’s law of motion. The system was assumed to be in an overdamped regime (mä = 0): (1) η⋅r˙i=Fi,i−1+Fi,i+1+FiB where ri was the position vector of the node i in three-dimensional coordinates (x,y,z); η was the drag coefficient for the fiber node in the blood serum; Fi,i–1 = Kr ∙ (∆ri,i–1 − L) and Fi,i+1 = Kr ∙ (∆ri,i+1 − L) were the vector forces which acted on the node i from the node i-1 or i+1 according to the Hook’s law due to the spring elongation or shortening (∆ri,i–1 and ∆ri,i+1 was the distance between nodes i and i-1 or the nodes i and i+1; Kr was the fiber spring constant); FiB was the Brownian force which acted on the node i due to thermal fluctuations. To solve the equations for the nodes in solution, the system of equations was discretized and the Heun’s method was used. According to this method, the evolution of coordinates in time for the differential equation (2) x˙=f(x,t) was calculated using a two-step algorithm: (3) x~n+1=xn+Δt⋅f(xn,tn) (4) xn+1=xn+Δt2⋅{f(xn,tn)+f(x~n+1,tn+1)} where ∆t was the time step used in calculation and discretization scheme; tn+1 = tn + ∆t; xn was coordinate at the tn moment of time.

Thus, the evolution of the node i coordinates in time was calculated according to a two-step process (in a vector form): (5) r~in+1=rin+Δtη⋅(Fi,i−1(rin,ri−1n)+Fi,i+1(rin,ri+1n)+Fin,B) (6) rin+1=rin+Δt2η⋅(Fi,i−1(rin,ri−1n)+Fi,i+1(rin,ri+1n)+Fin,B+Fi,i−1(r~in+1,r~i−1n+1)+Fi,i+1(r~in+1,r~i+1n+1)+F~in+1,B) The Brownian force in three dimensions was calculated as described previously.24 Briefly, the mean square displacement of a particle over a single time step was ⟨x2⟩=6kbTηdt, and the discretized Brownian force acting on the node i was Fin,B=εi⋅6kbTηdt, where εi was sampled from the standard normal distribution.

At the initial time point, the node 0 was bound to the integrin receptor on the bulb surface (top view, supplementary Fig. 3B). The distance of the binding point from the bulb top was D. The height of the binding point from the bulb surface was equal to the integrin αIIbβ3 molecule length (H = 20 nm) in its elongated conformation.42 The equations for the node 0 swirling and downstream motion were: (7) x0n+1=(R+H)⋅cos⁡(Rate_swirl⋅Δt+φn) (8) y0n+1=(R+H)⋅sin⁡(Rate_swirl⋅Δt+φn) (9) φn+1=Rate_swirl⋅Δt+φn (10) z0n+1=z0n−Rate⋅Δt When the node i entered the light-magenta zone (supplementary Fig. 3B), and its distance from the bulb surface became smaller than H, it bound the integrin molecule and its equations of motion became similar to the node 0 equations (7) – (10).

The integrin αIIbβ3 surface density on the bulb surface was relatively high (the distance between the neighboring molecules was approximately 15 nm). Given such a high density, we assumed that when the node became close enough to the bulb, it immediately encountered the integrin molecule and bound it. Thus, when the distance between the node i and the bulb surface became smaller than H, node i bound to the integrin molecule and the node’s equations of motion became similar to the node 0 equations (7) – (10). The binding was assumed to be irreversible.

The model parameters were listed in the Table 1 below. The model equations were solved using Python 3.11.